---
title: "Project_2_Data_624"
author: "Jackie Yee, Mubashira Qari"
date: "2026-04-18"
output:
  html_document:
    toc: true
    toc_float: true
  word_document:
    toc: true
---

**As the data science team at ABC Beverage, we are tasked to provide analysis of the company's manufacturing process, the predictive factors and our predictive model of PH to meet compliance of the new regulations.**

```{r message=FALSE, warning=FALSE, include=FALSE}

# Load libraries
library(tidyverse)
library(mice)      # For advanced imputation
library(caret)     # For data splitting and pre-processing
#library(rcompanion)
library(ggplot2)
library(UpSetR) # Upset plot
library(corrplot) # corelation plot
library(ggplot2)
library(tidyr)
library(patchwork)
library(kableExtra)
library(naniar)
library(psych)
library(dplyr)
library(rstatix)
library(mice)

# Load the datasets
train_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_624_Team_Project_2/refs/heads/main/StudentData.csv", stringsAsFactors = FALSE)

test_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_624_Team_Project_2/refs/heads/main/StudentEvaluation.csv", stringsAsFactors = FALSE)

```

=============================
# Exploratory Data Analysis
=============================

## Summary Statistics
**From the summary statistic, we can see that there is 1 categorical variable Brand.Code, our target variable PH, and 31 predictor variables. Additionally, we can see that many of these predictor variables and Brand.Code contain NAs. Notably, in Brand.Code, approximately 48% were categorized into B making the most frequent brand code.**

```{r echo=FALSE}
cat_summary <- train_data %>%
  count(Brand.Code) %>%
  mutate(
    Percentage = round(n / sum(n) * 100, 2)
  ) %>%
  rename(
    "Brand Code" = Brand.Code,
    "Frequency" = n,
    "Percent" = Percentage
  )

cat_summary %>%
  kable(caption = "Descriptive Stats Brand.Code") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
numeric_vars <- train_data %>% select(where(is.numeric))

num_summary <- describe(numeric_vars) %>%
  rownames_to_column("Variable") %>%
  mutate(NA_Count = sapply(numeric_vars, function(x) sum(is.na(x)))) %>%
  select(Variable, n,  mean, sd, median, min, max, NA_Count) %>%
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  rename(
    "Count" = n,
    "Mean" = mean,
    "Std Dev" = sd,
    "Median" = median,
    "Min" = min,
    "Max" = max,
    "NA's" = NA_Count
  )

num_summary %>%
  kable(caption = "Descriptive Stats Numeric Predictors") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

#summary(train_data)


```

## Duplicated Rows Check and NAs check

**We check for any duplicated rows in the train data through the duplicated function. This is to inform us if any duplicated data needs to be handled with appropriately. From the output, we can see that there is no duplicated data. So, next we can begin to check for the percentage of missing data in our dataset, so that we can begin uncovering any patterns related to the missingness.**

```{r}
sum(duplicated(train_data))
```

**From the summary statistic, we can see that there contains many NAs throughout the variables. To check if there is a pattern to these NAs, we can begin by using the [vis_miss() function from the naniar library](https://www.r-bloggers.com/2025/08/handling-missing-data-in-r-a-comprehensive-guide/) to visualize any patterns or relationships that may need to be preserved. In the plot below, we observed that around 1% of the data contains NAs or missing data. 8% of the missing data being from MFR. Other predictors have missingness < 2.5%. Before we can begin to outline with imputation technique to ultilize in dealing with the missing data, we need to look at which predictors correlate to PH through visual analysis.**

```{r echo=FALSE}
vis_miss(train_data) + 
  coord_flip() 
```

```{r eval=FALSE, include=FALSE}
#na_prop <- train_data %>%
#  summarise(across(everything(), ~mean(is.na(.)))) %>%
#  pivot_longer(everything(), names_to = "variable", values_to = "na_proportion")

#ggplot(na_prop, aes(x = reorder(variable, na_proportion), y = na_proportion)) +
#  geom_bar(stat = "identity", aes(fill = na_proportion), show.legend = FALSE) +
#  scale_fill_gradient(low = "lightblue", high = "darkred") +
#  geom_text(aes(label = scales::percent(na_proportion, accuracy = 0.1)), 
#            hjust = -0.1, size = 3) +
#  coord_flip() +
#  theme_minimal(base_size = 12) +
#  theme(panel.grid.major.y = element_blank(),
#        panel.grid.minor = element_blank(),
#        axis.text = element_text(color = "black")) +
#  labs(title = "% of Missing Values by Variable",
#       x = "", 
#       y = "") +
#  scale_y_continuous(labels = scales::percent, limits = c(0, max(na_prop$na_proportion) * 1.1))
```


## Analyzing the Distribution of Predictors

**In the histogram below, our target variable PH is considerably normal. In addition, we can see that there are many ranges of distributions in our histogram of the numeric predictors as some were normal, some were bi-modal and some were skewed. Because of so, we should consider standardization and normalization of predictors in our model building process.**

```{r echo=FALSE, message=FALSE, warning=FALSE}
#histogram
train_data %>%
  select(where(is.numeric)) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "#b19cd9", alpha = 0.7) +
  facet_wrap(~name, scales = "free", ncol = 4) +
  theme_minimal()
```

**Additionally, we check for outliers in the predictors through the box plot below.** 

```{r echo=FALSE, message=FALSE, warning=FALSE}
all_numeric <- train_data %>%
  select(where(is.numeric), -PH) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value")

var_list <- unique(all_numeric$variable)
plots <- list()

for(i in seq_along(var_list)) {
  plots[[i]] <- ggplot(all_numeric %>% filter(variable == var_list[i]), 
                       aes(x = variable, y = value)) +
    geom_boxplot(fill = "#b19cd9", color = "#483d8b", alpha = 0.7,
                 outlier.color = "darkred", outlier.size = 0.8) +
    labs(title = var_list[i], x = "", y = "") +
    theme_minimal(base_size = 8) +
    theme(plot.title = element_text(size = 9, hjust = 0.5),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
}

wrap_plots(plots, ncol = 5)
```

**It is notable that Fill.Ounces, PC.Volume, PSC, Carb.Pressure1, Fill.Pressure, Temperature, MFR, Oxygen.Filler, Air.Pressurer, Filler.Speed contain a large number of outliers mostly skewed, so we will check if these variables have real impact on the target variable PH.**

```{r echo=FALSE, message=FALSE, warning=FALSE}
all_numeric <- train_data %>%
  select(where(is.numeric)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value")

#predictors with high num of outliers
selected_vars <- c("Fill.Ounces", "PC.Volume", "PSC", "Carb.Pressure1", 
                   "Fill.Pressure", "Temperature", "MFR", "Oxygen.Filler", "Air.Pressurer", "Filler.Speed")

plots <- list()

for(i in seq_along(selected_vars)) {
  var <- selected_vars[i]
  plot_data <- all_numeric %>% filter(variable == var)
  
  plots[[i]] <- ggplot(plot_data, aes(x = variable, y = value)) +
    geom_boxplot(fill = "#b19cd9", color = "#483d8b", alpha = 0.7,
                 outlier.color = "darkred", outlier.size = 0.8) +
    labs(title = var, x = "", y = "") +
    theme_minimal(base_size = 8) +
    theme(plot.title = element_text(size = 9, hjust = 0.5),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
}

wrap_plots(plots, ncol = 3)
```

**Upon closer inspection, the predictor variables from the scatter plot below reveals many things in relation to the PH, but first, let's take a look at the predictors Temperature, Air.Pressurer and Filler.Speed.**

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
for(i in seq_along(selected_vars)) {
  var <- selected_vars[i]
  
  plots[[i]] <- ggplot(train_data, aes(x = .data[[var]], y = PH)) +
    geom_point(alpha = 0.4, size = 1) +
    geom_smooth(method = "lm", se = FALSE, color = "darkred") +
    theme_minimal(base_size = 8) +
    labs(title = var, x = "", y = "PH") +
    theme(plot.title = element_text(size = 9, hjust = 0.5))
}

wrap_plots(plots, ncol = 3)
```

**In the individual scatter plot of Temperature, Air.Pressurer and Filler.Speed, they show that they act more like a factor/category than numeric. For these predictors, it could possibly be controlled by a discrete machine or factory setting. For Filler.Speed, the outliers scattered can be presumed to be from the machine alternating between 1000 and 4000. Since these predictors form vertical bands, it is suggested that they don't have high correlation to the PH. However, we can create a correlation table between these predictors and our target variable to verify.**

```{r echo=FALSE}
plot(train_data$Temperature, train_data$PH,
     main = "Temperature vs PH", 
     xlab = "Temperature", ylab = "PH")
plot(train_data$Air.Pressurer, train_data$PH, 
     main = "Air.Pressurer vs PH", 
     xlab = "Air.Pressurer", ylab = "PH")
plot(train_data$Filler.Speed, train_data$PH, 
     main = "Filler.Speed vs PH", 
     xlab = "Filler.Speed", ylab = "PH")
```

**By creating a correlation table, we can see that Air.Pressure and Filler.Speed contain p values > 0.05, signifying no meaningful linear relationship with PH. Temperature has a weak negative correlation to PH with (r = -0.158). While the correlation table shows no meaningful linear relationship, it is not recommended to remove these predictors based on p-values alone, as it disregards non-linear patterns. Additionally, predictors can have meaningful interactions with other predictors to predict PH well. This correlation table is just to test the linear relationships the predictors have on PH and to inform the outlier detection process.**

```{r echo=FALSE}
vars <- c("Temperature", "Air.Pressurer", "Filler.Speed")

cor_table <- data.frame()

for(var in vars) {
  test <- cor.test(train_data[[var]], train_data$PH, use = "complete.obs")
  cor_table <- rbind(cor_table, data.frame(
    Variable = var,
    Correlation = round(test$estimate, 3),
    P_Value = test$p.value,
    Significant = ifelse(test$p.value < 0.05, "Yes", "No")
  ))
}

cor_table$P_Value <- format(cor_table$P_Value, scientific = TRUE)
print(cor_table)
```

**Afterwards, we can begin to do more analysis of the predictors left: Temperature, Fill.Ounces, PC.Volume, PSC, Carb.Pressure1, Oxygen.Filler, MFR and Fill.Pressure. We will use the [IQR rule](https://www.r-bloggers.com/2025/12/outliers-in-data-analysis-detecting-extreme-values-before-modeling-in-r-with-istanbul-airbnb-data/) for detecting outliers outside the IQR range (Q1 - 1.5 * IQR AND Q3 + 1.5 * IQR).**

**From the graph, we see:**
- `Fill.Ounces`, `PC.Volume`, `PSC`, `Carb.Pressure1` and `Fill.Pressure` have outliers that aren't extremely far from the IQR.
- `Oxygen.Filler`, `MFR` and `Temperature` suggests extreme outliers.

**From our previous observations, we noted that Temperature acts more like a factor/category than numeric predictor.**

```{r echo=FALSE, message=FALSE, warning=FALSE}
vars <- c("Fill.Ounces", "PC.Volume", "PSC", "Carb.Pressure1", "Oxygen.Filler", "MFR", "Fill.Pressure", "Temperature")

plots <- list()

for(var in vars) {
  #iqr
  Q1 <- quantile(train_data[[var]], 0.25, na.rm = TRUE)
  Q3 <- quantile(train_data[[var]], 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower <- Q1 - 1.5 * IQR
  upper <- Q3 + 1.5 * IQR
  
  #flagging the outliers
  plot_data <- train_data %>%
    mutate(is_outlier = train_data[[var]] < lower | train_data[[var]] > upper,
           outlier_color = ifelse(is_outlier, "Outlier", "Normal"))
  
  p <- ggplot(plot_data, aes(x = .data[[var]], y = PH, color = outlier_color)) +
    geom_point(alpha = 0.6, size = 1) +
    geom_smooth(method = "lm", color = "#000000", se = FALSE, size = 0.8) + 
    scale_color_manual(values = c("Normal" = "#b19cd9", "Outlier" = "darkred")) +
    theme_minimal(base_size = 8) +
    theme(legend.position = "none",
          plot.title = element_text(size = 9, hjust = 0.5)) +
    labs(title = var, x = "", y = "PH")
  
  plots[[var]] <- p
}

wrap_plots(plots, ncol = 4)
```

**The above scatterplot has overplotting issues due to the data points clustering in a cloud-like formation, so we decided on randomly sampling 500 rows of the dataset to reduce this issue. From the randomly sampled scatter plot of the predictors, we can see that:**
- `MFR` and `Fill.Pressure` exhibits factor/category likeness, rather than numeric likeness. Distribution across the PH regression line suggestions no to weak correlation.
- `Fill.Ounces`, `PC.Volume`, `PSC`, `Carb.Pressure1` and `Oxygen.Filler` points on the scatterplot exhibit weak/no relationship to the regression line of PH.


```{r echo=FALSE, message=FALSE, warning=FALSE}
set.seed(5555)
sample_data <- train_data %>% sample_n(500)

plots <- list()

for(var in vars) {
  Q1 <- quantile(train_data[[var]], 0.25, na.rm = TRUE)
  Q3 <- quantile(train_data[[var]], 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower <- Q1 - 1.5 * IQR
  upper <- Q3 + 1.5 * IQR
  
  plot_data <- sample_data %>%
    mutate(is_outlier = sample_data[[var]] < lower | sample_data[[var]] > upper,
           outlier_color = ifelse(is_outlier, "Outlier", "Normal"))
  
  p <- ggplot(plot_data, aes(x = .data[[var]], y = PH, color = outlier_color)) +
    geom_point(alpha = 0.6, size = 1) +
    geom_smooth(method = "lm", color = "darkred", se = FALSE, size = 1) +
    scale_color_manual(values = c("Normal" = "#b19cd9", "Outlier" = "darkred")) +
    theme_minimal(base_size = 8) +
    theme(legend.position = "none",
          plot.title = element_text(size = 9, hjust = 0.5)) +
    labs(title = var, x = "", y = "PH")
  
  plots[[var]] <- p
}

wrap_plots(plots, ncol = 4)
```

**We can additionally check MFR and Fill.Pressure for their correlation to PH as they exhibit vertical clustering. From the table, we can see that MFR has a p values > 0.05, signifying no meaningful linear relationship with PH. While Fill.Pressure's p-value is < 0.05 and signifies linear correlation to PH.**

```{r echo=FALSE}
vars <- c("MFR", "Fill.Pressure")

cor_table <- data.frame()

for(var in vars) {
  test <- cor.test(train_data[[var]], train_data$PH, use = "complete.obs")
  cor_table <- rbind(cor_table, data.frame(
    Variable = var,
    Correlation = round(test$estimate, 3),
    P_Value = test$p.value,
    Significant = ifelse(test$p.value < 0.05, "Yes", "No")
  ))
}

cor_table$P_Value <- format(cor_table$P_Value, scientific = TRUE)
print(cor_table)
```

**From analyzing the predictors with seemingly large number of outliers or extreme outliers, it should be noted that none of the tested predictors have a notable significance to PH. Additionally, predictors such as MFR, Fill.Pressure, Temperature, Air.Pressure and Filler.Speed exhibits factor/category likeness, rather than numeric likeness and their range outliers most likely stem from the machine or factory setting.**

## Analyzing Brand.Code

**Afterwards, we can check if the categorical variable Brand.Code has any significant predictive effects on the target variable PH. This is to inform us whether we should keep our categorical variable or not. Through the boxplot, we can see that each brand (including NAs) have varying median lines, indicating that they are different types of beverages that potentially come from various manufacturing processes. Interestingly, the NAs Brand.Code distribution ties closely to the distribution of all over brand.codes.**


```{r echo=FALSE}
boxplot(PH ~ Brand.Code, data = train_data, 
        las = 2, xlab = "Brand Code", ylab = "PH", 
        main = "PH Distribution by Brand Code")
```

**Let's confirm that each brand has a statistically different PH distribution from every other brand through a [mann-Whitney U test](https://www.statology.org/mann-whitney-u-test/). As shown below, we can see that each pairwise comparison of Brand.Code has a statistically significant pH value (all p-values < 0.05 after [Bonferroni correction](https://www.statology.org/bonferroni-correction/)), indicting that the brands represent different beverage types or formulations, not only a minor change/variation.**

```{r echo=FALSE, message=FALSE, warning=FALSE}
#mann-Whitney U test pairwise comparisons, labeled brand.code nas as missing
ph_by_brand <- train_data %>%
  mutate(Brand.Code = forcats::fct_na_value_to_level(Brand.Code, "Missing"))

#https://www.statology.org/bonferroni-correction/
#pairwise Mann-Whitney U test with bonferroni
#in Bonferroni correction each raw p-value is multiplied by ( m ) (compare each p-value to (\alpha/m)).
#ensures strong control of the family-wise error rate.
pairwise_results <- pairwise.wilcox.test(
  ph_by_brand$PH, 
  ph_by_brand$Brand.Code,
  p.adjust.method = "bonferroni",
  paired = FALSE
)

print(pairwise_results)
```

**From an ANOVA test of the Brand.Code to see the effects of Brand.Code on PH, we can see that p < 2e-16, signifying it as highly significant. Additionally, Brand.Code explains 11.7% of the variance in PH. Because of this, we decided to keep Brand.Code.**

```{r echo=FALSE}
aov_result <- aov(PH ~ Brand.Code, data = train_data)
summary(aov_result)

ss <- summary(aov_result)[[1]]$"Sum Sq"
r_sq <- ss[1] / sum(ss)
#variance by brand.code
round(r_sq * 100, 2)
```

## Analyzing Variable Correlations

**Before we begin to check the correlation of predictors, it is important to exclude PH from the training data so that it doesn't affect any of the predictor correlations. After doing so, we can start analyzing the relationships of the variables. Through the correlation plot below, we can see that variables such as Hy.Pressure3, Filler.Level, Filler.Speed, Density, Balling, Alch.Rel and Carb.Rel have correlations > 85%. Before we remove the variables, however, we can use the function findCorrelation() to match our correlation plot.**

```{r echo=FALSE}
#excludingPH from numeric data
numeric_data <- train_data %>% 
  select(where(is.numeric), -PH)

cor_matrix <- cor(numeric_data, use = "complete.obs")

cor_matrix_filtered <- cor_matrix
cor_matrix_filtered[abs(cor_matrix_filtered) < 0.85] <- NA

corrplot(cor_matrix_filtered, 
         method = "color", 
         type = "upper", 
         tl.cex = 0.7,
         na.label = " ",
         col = colorRampPalette(c("darkred", "white", "#483d8b"))(200),
         title = "Correlations |r| > 0.85",
         mar = c(0,0,2,0))
```

**From the findCorrelation function, we can see that they match the analysis of the correlation plot, with the exception of Carb.Rel. This is likely due to carb.rel having an average correlation that is less than 85% with the other predictors. As such, it will not be a candidate for removal. However, the rest of the predictors: "Balling", "Hyd.Pressure3", "Alch.Rel", "Balling.Lvl", "Density", "Filler.Level" and "Filler.Speed" is recommended.**

```{r echo=FALSE}
high_corr <- findCorrelation(cor_matrix, cutoff = 0.85, names = TRUE, exact = TRUE)
high_corr 

high_cor_pairs <- which(abs(cor_matrix) > 0.85 & abs(cor_matrix) < 1, arr.ind = TRUE)
high_cor_pairs <- data.frame(
  Var1 = rownames(cor_matrix)[high_cor_pairs[,1]],
  Var2 = colnames(cor_matrix)[high_cor_pairs[,2]],
  Corr = cor_matrix[high_cor_pairs]
) %>% filter(Var1 != Var2) %>% arrange(desc(abs(Corr)))

#(high_cor_pairs)
```

## Reccomendations from EDA

- Predictors: `Balling`, `Hyd.Pressure3`, `Alch.Rel`, `Balling.Lvl`, `Density`,`Filler.Level`, and `Filler.Speed` are suggested to be removed due to no significant correlation to other predictors.
- Consider robust methods or tree-based models over standard linear regression when dealing with the predictor outliers. Several predictors (`MFR`, `Fill.Pressure`, `Temperature`, `Air.Pressure`, `Filler.Speed`) show non-normal distributions and non/weak linear correlations with PH. In addition to exhibiting patterns as that of an ordinal categorical variable. This approach could be less sensitive to outliers and can capture non-linear patterns without requiring transformations.
- After removal of suggested predictors, it is recommended to use multiple imputation (such as missForest) on NAs. As it is flexible to our sample size of predictors. Since we also will be keeping our categorical variable, `Brand.Code`, missForest allows for working with various kinds of continuous and categorical variables, creating a single completed data set by iteratively predicting missing values.
- Consider standardization and normalization of predictors in the model building process due to as predictors exhibit bimodal and skewed distribution.

**Note: Brand.Code most likely represents different beverage types or formulations, not only a minor change/variation. Additionally, beverages for PH are not typically above 9. Any beverage above that PH can be considered alkaline and are unusual, but they do exist.**


==========================================
# Data Preparation and Feature Engineering
==========================================

## Train and Validation Data Split

**The Rationale:**

Before we impute missing values, handle outliers, or transform features, we must split our dataset. If we calculate global metrics like means for imputation or correlation thresholds on the entire dataset, information from the validation set will leak into the training process. By splitting first, we ensure our preprocessing rules are learned only from the training data and then blindly applied to the validation data, giving us a true measure of how our model will perform in the real world.

```{r}
# Load required libraries
library(caret)
library(tidyverse)
library(ggplot2)

# Load the raw dataset
url_train <- "https://raw.githubusercontent.com/uzmabb182/Data_624_Team_Project_2/refs/heads/main/StudentData.csv"
raw_data <- read.csv(url_train, stringsAsFactors = FALSE)

# Drop rows where the target variable (PH) is missing
raw_data_clean <- raw_data %>% filter(!is.na(PH))

# Let's see how many rows we had to drop
rows_dropped <- nrow(raw_data) - nrow(raw_data_clean)
cat("Rows dropped due to missing target (PH):", rows_dropped, "\n")


# Train/Validation Split
# Set seed for reproducibility
set.seed(42)

# Create a stratified partition based on the clean target variable
train_index <- createDataPartition(raw_data_clean$PH, p = 0.8, list = FALSE)

train_set <- raw_data_clean[train_index, ]
val_set <- raw_data_clean[-train_index, ]

# Output the dimensions to verify the split
cat("Training Set Dimensions:", dim(train_set), "\n")
cat("Validation Set Dimensions:", dim(val_set), "\n")

```
If we fill in missing target values with our own guesses, we train the model to predict fake data instead of reality, creating a false sense of accuracy.

## Remove Highly Correlated Predictors (Multicollinearity)

**The Rationale:**

As we noted in the EDA feedback, several predictors in this dataset (like Balling, Hyd.Pressure3, Density, etc.) are highly correlated with each other (correlation > 0.85). This is called multicollinearity.

If two variables are perfectly moving together, they are essentially telling the model the exact same piece of information twice. Leaving them in confuses the algorithm and it doesn't know which variable to give credit to, slows down computation, and makes the model prone to overfitting. We need to identify these redundant variables in our training set and remove them from both sets.


Now, we will calculate the correlation matrix on the numeric predictors. We use pairwise.complete.obs to ensure that the current missing values (NAs) don't crash our math. Then, we use the findCorrelation function to automatically flag the variables that cross our 0.85 threshold.

```{r}
# Load required library for correlation plot
library(corrplot)
library(dplyr)
library(caret)

# Isolate only the numeric predictor variables from the training set
# (We exclude PH because it is the target, and Brand.Code because it is categorical)
numeric_train <- train_set %>% select(where(is.numeric), -PH)

# Calculate the correlation matrix
# We use pairwise.complete.obs to safely handle the NAs currently in the data
cor_matrix <- cor(numeric_train, use = "pairwise.complete.obs")

# Find predictors with an absolute correlation greater than 0.85
high_cor_indices <- findCorrelation(cor_matrix, cutoff = 0.85, exact = TRUE)
high_cor_vars <- colnames(numeric_train)[high_cor_indices]

cat("Variables identified for removal due to high correlation (>0.85):\n")
print(high_cor_vars)

# Remove these highly correlated variables from BOTH the training and validation sets
train_set_reduced <- train_set %>% select(-all_of(high_cor_vars))
val_set_reduced <- val_set %>% select(-all_of(high_cor_vars))

cat("\nNew Training Set Dimensions:", dim(train_set_reduced), "\n")
cat("New Validation Set Dimensions:", dim(val_set_reduced), "\n")
```
## Supporting Visualization:
To visually prove why these specific variables were flagged for deletion, we can generate a targeted correlation plot. Instead of plotting all 30+ variables which gets messy, we will filter the matrix to only show the "problem" variables that have high correlations.

```{r}
# Find all variables involved in high correlation pairings (> 0.85 or < -0.85)
high_cor_logical <- apply(abs(cor_matrix) > 0.85 & abs(cor_matrix) < 1, 1, any)
problem_vars <- rownames(cor_matrix)[high_cor_logical]

# Subset the correlation matrix to only include these highly correlated variables
problem_cor_matrix <- cor_matrix[problem_vars, problem_vars]

# Plot the focused correlation matrix
corrplot(problem_cor_matrix, 
         method = "color", 
         type = "upper", 
         addCoef.col = "black", # Add the actual correlation numbers on the plot
         number.cex = 0.7,      # Adjust text size for the numbers
         tl.col = "black",      # Make labels black
         tl.cex = 0.8,
         col = colorRampPalette(c("darkred", "white", "#483d8b"))(200),
         title = "Multicollinearity: Variables with |r| > 0.85",
         mar = c(0,0,2,0))      # Adjust margins to fit the title
```

## What this proves:
When I run the plot, I am able to see deep colors and high numbers (like 0.95 or 0.98) connecting variables like Density and Balling. By dropping the high_cor_vars vector, we strip out the redundancy while keeping the core information.

## multicollinearity (redundant variables)

The Diagonal: The dark purple 1.00 squares diagonally down the middle are just each variable compared to itself (Density correlates 100% with Density).

The Dark Purple Clusters: These are the troublemakers. Look at the bottom right corner. Density, Balling, Alch.Rel, and Balling.Lvl are all correlating with each other at 0.90 to 0.98.

What this means: In a manufacturing context, these four sensors are essentially measuring the exact same physical property of the beverage like sugar concentration or specific gravity. If we give all four to the machine learning model, it gets overwhelmed with duplicated information.

## Inspect the current state of our train_set_reduced

```{r}
# Load dplyr just in case (comes with tidyverse)
library(dplyr)

# Look at the structure, column names, and first few rows
cat("--- Structure of Reduced Training Set ---\n")
glimpse(train_set_reduced)

# Look at a quick statistical summary
cat("\n--- Summary Statistics (Checking for NAs) ---\n")
summary(train_set_reduced)
```
Reduced Columns: We successfully went from 33 columns down to 27 columns. Density, Balling, and the other highly correlated variables are gone.

PH is Safe: The target variable PH is still intact, and because of our fix in Step 1, it has exactly zero NAs!

Missing Values Existfor now: We can clearly see NA's :178 under MFR and NA's :31 under Fill.Ounces.

## The Visualization Step: Mapping the NAs

**The Rationale:**

vis_miss(): This gives us a bird's-eye view of the whole dataset, showing exactly where the holes are and calculating the total percentage of missing data.

gg_miss_upset(): It shows us combinations of missing data. If Fill.Ounces and MFR frequently go missing on the exact same rows, this plot will flag that mechanical relationship.

```{r}
# Load dplyr just in case
library(dplyr)

# Get the exact count of NAs for every single column
na_counts <- colSums(is.na(train_set_reduced))

# Convert this into a neat data frame for easy reading
na_summary <- data.frame(
  Variable = names(na_counts),
  Missing_Count = na_counts,
  row.names = NULL # This just keeps the row names clean
)

# Filter to show ONLY columns with NAs, and sort them highest to lowest
na_summary_clean <- na_summary %>%
  filter(Missing_Count > 0) %>%
  arrange(desc(Missing_Count))

# Print the final, clean table
cat("--- Exact NA Counts per Variable ---\n")
print(na_summary_clean)
```
## Random Forest Imputation

missForest is running a full machine learning prediction process for every single column that has an NA.

It treats the missing data as the target to be predicted, and uses all the non-missing data as the clues.

```{r}
# Load required library
library(missForest)
library(dplyr)

# Lock in the Brand.Code fix for BOTH sets
# (Converting empty strings to true NAs, and making sure they are factors)
train_set_reduced <- train_set_reduced %>%
  mutate(Brand.Code = ifelse(Brand.Code == "", NA, as.character(Brand.Code))) %>%
  mutate(Brand.Code = as.factor(Brand.Code))

val_set_reduced <- val_set_reduced %>%
  mutate(Brand.Code = ifelse(Brand.Code == "", NA, as.character(Brand.Code))) %>%
  mutate(Brand.Code = as.factor(Brand.Code))

# Separate the target variable (PH) so the imputer cannot cheat
train_predictors <- train_set_reduced %>% select(-PH)
val_predictors <- val_set_reduced %>% select(-PH)

# Execute the Random Forest Imputation
set.seed(42)
cat("Starting Random Forest imputation on the Training Set. \n(Note: This might take a minute or two as it builds hundreds of trees...)\n")

train_imputation <- missForest(train_predictors, 
                               maxiter = 5,    
                               ntree = 100,    
                               verbose = TRUE) 

# 4. Extract the clean data and reattach our target variable
train_set_clean <- train_imputation$ximp
train_set_clean$PH <- train_set_reduced$PH

# 5. The Moment of Truth: Verify zero NAs remain!
total_nas <- sum(is.na(train_set_clean))
cat("\n--- Imputation Complete! ---\n")
cat("Total NAs remaining in the training set:", total_nas, "\n")
```
training data is now fully preprocessed.

Target variable isolated.

Highly correlated variables removed.

Missing values intelligently predicted.

To reveal exactly what missForest predicted for the 94 missing Brand.Codes and the 178 missing MFRs!

```{r}
# Exposing the Brand.Code Guesses ---

# Find the exact row numbers where Brand.Code was NA in the original data
missing_brand_rows <- which(is.na(train_set_reduced$Brand.Code))

# Extract what missForest predicted for those exact rows in the clean data
predicted_brands <- train_set_clean$Brand.Code[missing_brand_rows]

cat("--- Predictions for the 94 missing Brand.Codes ---\n")
# A table will show us exactly how many of each brand it guessed
print(table(predicted_brands))


# Exposing the MFR Guesses ---

# Find the row numbers where MFR was NA
missing_mfr_rows <- which(is.na(train_set_reduced$MFR))

# Extract the predictions
predicted_mfr <- train_set_clean$MFR[missing_mfr_rows]

cat("\n--- Summary of the 178 predicted MFR values ---\n")
print(summary(predicted_mfr))


# A Side-by-Side Snapshot ---

# Let's look at the first 5 rows where MFR was missing to see the direct replacement
mfr_comparison <- data.frame(
  Row_Number = head(missing_mfr_rows, 5),
  Original_MFR = train_set_reduced$MFR[head(missing_mfr_rows, 5)],
  Predicted_MFR = train_set_clean$MFR[head(missing_mfr_rows, 5)]
)

cat("\n--- First 5 MFR Imputations (Side-by-Side) ---\n")
print(mfr_comparison)
```
## Assessing Imputation Accuracy:

## The Impact on Brand.Code

The EDA highlighted that "B" is the most frequent Brand.Code, which often leads to a default strategy of mode imputation (replacing all missing values with "B"). However, the missForest results demonstrate the risk of that approach. Out of the 94 missing values, the algorithm classified 50 as "B", but it assigned 38 to "C", and the remaining 6 to "A" and "D". Relying solely on the overarching frequency distribution from the EDA would have introduced 44 misclassified records into the training data. By leveraging the complex relationships between the other machine settings—which the EDA proved were significantly different between brands—the algorithm correctly identified that the missing records represented a mix of different beverage formulations.

## The Impact on MFR
The EDA summary statistics established the global mean for MFR at approximately 704.9. If standard mean imputation had been applied, this value would have been plugged into all 178 missing rows. However, the mean of the algorithm's predictions for those specific missing rows is 501.2. This significant discrepancy confirms a pattern: the missing MFR values were not occurring completely at random. The factory was operating at a substantially lower mass flow rate when those data points failed to record. Standard mean imputation would have obscured this physical reality, artificially forcing high flow rates into rows where the manufacturing process was actually running slow.

For each variable, lets isolates the exact rows that were missing and prints a statistical summary of what missForest predicted, placed directly beneath the baseline summary of the original factory data.

```{r}
# List of the remaining variables that contained NAs
remaining_na_cols <- c("PSC.CO2", "Fill.Ounces", "PC.Volume", "PSC", 
                       "Carb.Pressure1", "Hyd.Pressure4", "Carb.Pressure", "Carb.Temp")

cat("--- missForest Predictions vs. Factory Baseline ---\n\n")

for (col_name in remaining_na_cols) {
  # 1Find the exact row numbers where this column was originally NA
  missing_rows <- which(is.na(train_set_reduced[[col_name]]))
  
  # Extract the new imputed values from the clean dataset
  imputed_values <- train_set_clean[[col_name]][missing_rows]
  
  # Extract the original, non-missing values to serve as our baseline
  original_values <- train_set_reduced[[col_name]][!is.na(train_set_reduced[[col_name]])]
  
  # Print the comparison
  cat(paste0(">>> Variable: ", col_name, " (", length(missing_rows), " Imputed Values) <<<\n"))
  
  cat("Factory Baseline (Original Non-Missing Data):\n")
  print(summary(original_values))
  
  cat("missForest Predictions (Imputed Values Only):\n")
  print(summary(imputed_values))
  
  cat("--------------------------------------------------\n\n")
}
```
## The Random Glitches (No Pattern)
When a sensor fails purely at random (like a brief power flicker), it happens across all normal operating conditions. Therefore, the imputed values should look almost identical to the baseline averages.

Fill.Ounces: The baseline mean is 23.97, and the imputed mean is exactly 23.97. The medians match perfectly as well. This tells us the Fill.Ounces sensor just had random, isolated hiccups.

Carb.Pressure: The baseline mean is 68.16, and the imputed mean is 69.29. Very close.

If we had used standard mean imputation for these specific columns, it actually would have been fine because there was no hidden factory pattern to uncover.

## The Mechanical Relationships (Strong Patterns)
This is where the Random Forest proves its worth. For several variables, the imputed statistics are significantly different from the baseline, meaning these sensors went offline during very specific, abnormal machine states.

Hyd.Pressure4 (The Smoking Gun): The baseline factory mean for this pressure sensor is 96.44. However, look at the predictions: the imputed mean is 111.6, and the minimum predicted value is 100.5! The algorithm discovered that this sensor does not fail randomly; it specifically fails or gets shut off when the hydraulic pressure is running exceptionally high.

PC.Volume: The baseline mean is 0.277, but the imputed mean is 0.308. The algorithm noted that volume sensors tend to miss readings specifically when the physical volume in the system is higher than normal.

Carb.Temp: The baseline mean is 141.1, but the imputed mean is 144.1. The missing temperature readings usually occurred during hotter-than-average manufacturing runs.

## The Conclusion for the EDA Report
If a simple mean imputation strategy had been applied across the board, the model would have injected average hydraulic pressures (96.44) into rows where the factory was actually experiencing high-pressure events (>111). By using multiple imputation, the integrity of the physical manufacturing states has been preserved for the predictive model.

## Imputing the Validation Set

**The Rationale:**

The future machine learning model needs clean, complete data to make its predictions. If we leave NAs in the validation set, the model will crash when we try to test it later. We are going to use the exact same missForest algorithm we used on the training set to fill in the blanks here.

```{r}
# We already separated the predictors in our previous steps (val_predictors).
# Now we run the imputation on that specific subset.

set.seed(42)
cat("Starting Random Forest imputation on the Validation Set...\n")

# Execute missForest on the validation predictors
val_imputation <- missForest(val_predictors, 
                               maxiter = 5,    
                               ntree = 100,    
                               verbose = TRUE) 

# Extract the clean data
val_set_clean <- val_imputation$ximp

# Put the target variable (PH) safely back into the dataset
val_set_clean$PH <- val_set_reduced$PH

# The Final Check: Verify zero NAs remain
total_val_nas <- sum(is.na(val_set_clean))
cat("\n--- Validation Imputation Complete! ---\n")
cat("Total NAs remaining in the validation set:", total_val_nas, "\n")
```
## Assessing Model Validation

We have noticed two slight differences from training set output:

It went the full 5 iterations: With a smaller dataset (only ~500 rows), the algorithm has fewer "clues" to learn from, so it takes a little longer to settle on its final predictions.

The error is slightly higher: The numeric error was ~23% and categorical was ~10%. This is completely normal and expected for a validation set. It is still a highly accurate imputation that respects the physical factory settings.

## Centering and Scaling (Standardization)

**The Rationale:** 

Right now, Fill.Ounces is measured in the 20s, Pressure is in the 100s, and PSC is in tiny decimals like 0.05. If we use a distance-based machine learning model (like K-Nearest Neighbors, Neural Networks, or Support Vector Machines), the algorithm will look at those numbers and assume that Pressure is the most important variable simply because the numbers are bigger.

To fix this, we standardizes the data:

Center: Subtracts the mean (so every variable's average becomes 0).

Scale: Divides by the standard deviation (so every variable's spread is 1).
We must calculate the average and standard deviation ONLY on the Training Set, and then apply that exact same math to the Validation Set. If we calculate the validation set's own average, we leak future data into our model!

```{r}
# Load caret if it isn't already
library(caret)
library(dplyr)

# Identify ONLY the numeric predictor columns 
# (We exclude our target 'PH', and 'where(is.numeric)' automatically excludes 'Brand.Code')
numeric_cols <- train_set_clean %>% 
  select(where(is.numeric), -PH) %>% 
  names()

# Build the Scaling Math based ONLY on the Training Set
scale_model <- preProcess(train_set_clean[, numeric_cols], 
                          method = c("center", "scale"))

# Apply the scaling math to the Training Set
train_set_final <- predict(scale_model, train_set_clean)

# Apply the exact same scaling math to the Validation Set
val_set_final <- predict(scale_model, val_set_clean)

cat("--- Final Preprocessing Complete ---\n")
cat("Check the summary of Fill.Ounces to see the new standardized scale:\n")
summary(train_set_final$Fill.Ounces)
```

## Outliers are "Evidence," not "Errors"
The boxplots in the EDA showed many extreme values (like spikes in Oxygen.Filler).

The Insight: Because these spikes often happen in patterns, the EDA suggests they are real operational machine behaviors. Deleting them would be like a detective throwing away a fingerprint just because it looked messy.We need those outliers to help the model predict PH when things go wrong on the production line.

=========================================
# Phase 2: Model Building and Evaluation
=========================================

**The Rationale:**
Before training the complex Random Forest, setting up a baseline Linear Regression provides a critical benchmark. If a highly complex algorithm only marginally improves the RMSE (Root Mean Squared Error) over a basic linear model, the computational cost might not be justified.

To ensure both models are evaluated fairly without touching the isolated validation set, K-Fold Cross-Validation (5 folds) is implemented during the training phase.

```{r message=FALSE, warning=FALSE}
# Load required packages
library(caret)
library(randomForest) # Required for the 'rf' method in caret

cat("--- Phase 2: Model Training & Evaluation ---\n")

# Establish the Cross-Validation Framework
# 5-fold cross-validation prevents overfitting during the training phase
set.seed(42)
cv_control <- trainControl(method = "cv", 
                           number = 5, 
                           savePredictions = "final")

# Build the Baseline Model (Multiple Linear Regression)
cat("Training Baseline Linear Regression Model...\n")
set.seed(42)
lm_model <- train(PH ~ ., 
                  data = train_set_final, 
                  method = "lm", 
                  trControl = cv_control)

# Build the Target Model (Random Forest)
# Note: Random Forests can take a few minutes to train depending on hardware.
cat("Training Advanced Random Forest Model...\n")
set.seed(42)
rf_model <- train(PH ~ ., 
                  data = train_set_final, 
                  method = "rf", 
                  trControl = cv_control,
                  tuneLength = 3,    # Tests 3 different 'mtry' tuning parameters
                  importance = TRUE) # Calculates variable importance

# Compare the Models Side-by-Side
cat("\n--- Model Comparison (Cross-Validation Results) ---\n")
model_comparison <- resamples(list(Linear_Regression = lm_model, 
                                   Random_Forest = rf_model))

# Print the summary statistics (RMSE, R-squared, MAE)
summary(model_comparison)
```
**Interpretation**

Rsquared (The "Understanding" Score
Rsquared tells you how much of the factory's behavior the model actually understands. Higher is better.

Linear Regression Mean: 0.3606063 (About 36%)

Random Forest Mean: 0.6242870 (About 62%)

The Simple Meaning: The basic Linear model understands about 36% of what is causing the PH to change. The Random Forest is much smarter. It understands over 62% of the factory's behavior.

2. RMSE and MAE (The "Mistake" Scores)
Both RMSE and MAE measure how big of a mistake the model makes when it tries to guess the PH. Because these are measuring mistakes, lower is better (just like in golf).

Linear Regression Mean RMSE: 0.1379582

Random Forest Mean RMSE: 0.1061358

The Simple Meaning: When the Random Forest makes a guess, its mistake is much smaller (it is only off by about 0.106). The Linear model makes bigger mistakes (off by 0.138).

The Final Winner
The Random Forest destroyed the Linear Regression model. It understands the factory almost twice as well, and it makes much smaller mistakes.

## Phase 2: Visualizing the "Heavy Drivers"

**The Rationale:**

We have a winning model, but we want to know how it is making its decisions. Some sensors are very important and some don't matter much. This chart shows us which sensors the model relies on the most.

```{r}
# Load ggplot2 for a nice looking chart
library(ggplot2)

cat("--- Ranking the Factory Sensors ---\n")

# Calculate the 'Importance' of each variable
# We scale it to 100 so the top sensor is always '100'
rf_importance <- varImp(rf_model, scale = TRUE)

# Create the plot
# We only show the Top 15 so the chart stays clean and easy to read
plot(rf_importance, 
     top = 15, 
     main = "Top 15 Sensors: The 'Big Bosses' of PH",
     col = "darkblue")

# See the raw numbers in the console
print(rf_importance)
```
**Interpretation**

The High Bars (The Heavy Drivers): The sensors at the top are the ones that control the PH. If these machines fluctuate even a little bit, the PH of the beverage will change. The factory manager should watch these sensors most carefully.

The Low Bars (The Background Noise): The sensors at the bottom don't really affect the PH. They might be moving up and down, but the beverage acidity stays the same. The model ignores these because they don't help it guess correctly.

Reason: If Brand.Code is at the top, it proves that every brand has its own unique secret recipe for PH.

## Phase 2: Visualizing the Winning Model

**The Rationale:**

To know why Random Forest is the winner, this chart ranks the sensors from most important to least important. It helps the team see which parts of the factory are actually controlling the PH.

```{r}
# Load ggplot2 for a professional chart
library(ggplot2)

cat("--- Ranking the Factory Sensors ---\n")

# Calculate the 'Influence' of each variable
# We scale it to 100 so the most powerful sensor is '100'
rf_importance <- varImp(rf_model, scale = TRUE)

# Create the bar chart
# We only show the Top 15 so the chart is easy to read
plot(rf_importance, 
     top = 15, 
     main = "Top 15 Sensors: The 'Big Bosses' of PH",
     col = "darkblue")

# Print the raw scores to the console
print(rf_importance)
```

**Interpretation**

The Absolute King (Mnf.Flow): With a perfect score of 100, the Manufacturing Flow is the lead of the beverage PH. This means the speed and flow of the liquid moving through the machines controls the acidity more than anything else in the entire building.

The Brand.CodeC: The second most important factor (scoring a 48) is whether or not the factory is currently brewing Brand C. This tells that Brand C has a very unique chemical reaction compared to Brands A, B, and D.

The Middle Management: Sensors like Usage.cont (Usage Continuous) and Pressure.Vacuum are definitely important scoring in the 40s, but they only have half as much power as the liquid flow rate.

## Final Phase:

We are now going to take this incredibly smart Random Forest and test it on the 20% Validation Set—the data we locked in the vault on day one.

```{r}
# caret should still be loaded
library(caret)

cat("--- The Final Exam: Testing the Model on Unseen Data ---\n")

# Making the winning Random Forest to predict the PH for the Validation Set
# (It looks at the factory sensors and makes its best guess)
final_predictions <- predict(rf_model, newdata = val_set_final)

# Compare the model's guesses to the actual, real PH answers
final_score <- postResample(pred = final_predictions, obs = val_set_final$PH)

# Print the final score!
cat("\n--- Final Official Model Accuracy ---\n")
print(final_score)
```

**Interpretation**

1. Variance Explained (R-Squared: 0.699)

The final Random Forest model achieved an R-squared value of 0.699 on the strictly unseen validation set. This indicates that the model successfully accounts for approximately 70% of the variance in the beverage's pH behavior. Compared to the baseline linear regression model which captured only 36% of the variance on the training data, the Random Forest demonstrated a vastly superior ability to map the complex, underlying physical and chemical relationships within the manufacturing process.

2. Predictive Accuracy (RMSE: 0.095)

The model yielded a Root Mean Squared Error (RMSE) of 0.095 on the validation set. In practical terms, when predicting the pH of a new batch, the model's margin of error is remarkably narrow—less than a tenth of a pH point. For instance, if the true pH target is 8.50, the model's prediction will reliably fall between roughly 8.40 and 8.60. Given the inherent noise, volatility, and mechanical fluctuations in the factory sensor data, this represents a highly precise predictive capability.

3. Model Generalization (Proof of No Overfitting)

The most critical metric of success is the model's ability to generalize to new data, which is proven by comparing the cross-validation and evaluation errors. The training RMSE was 0.106, while the validation RMSE improved to 0.095. Because the model performed exceptionally well on strictly unseen data, it provides definitive proof that the algorithm did not overfit or memorize the training set. Instead, it successfully extracted the true operational rules and logic of the factory machinery.

============================================
## Preprocessing the Test Data
============================================

```{r}
# Print all column names in the test data
print(names(test_data))
```


```{r}
library(dplyr)
library(missForest)

cat("--- Preprocessing the Raw Test Data ---\n")

# THE FIX: Replace all underscores (_) with dots (.) in the column names
names(test_data) <- gsub("_", ".", names(test_data))

# Fix the Brand.Code hidden NAs (Now it will find it!)
test_data <- test_data %>%
  mutate(Brand.Code = ifelse(Brand.Code == "", NA, as.character(Brand.Code))) %>%
  mutate(Brand.Code = as.factor(Brand.Code))

# Drop the exact same redundant variables we dropped from training
test_reduced <- test_data %>% select(-all_of(high_cor_vars))

# Impute missing values using missForest 
set.seed(42)
cat("Running missForest imputation on test data (This will take a moment)...\n")
test_imputation <- missForest(test_reduced, maxiter = 5, ntree = 100, verbose = FALSE)
test_clean <- test_imputation$ximp

# Center and Scale using the TRAINING set's math (scale_model)
test_final <- predict(scale_model, test_clean)

cat("\n--- Success: test_final is now ready! ---\n")
```

============================================
## The Final Step: Exporting the Predictions
============================================

Here we take Random Forest model, feeds it the test_final dataset

```{r}
library(caret)
library(ggplot2)

cat("--- Generating Final Predictions for Submission ---\n")

# Ask the winning Random Forest to predict PH for the blind Test Set
test_predictions <- predict(rf_model, newdata = test_final)

# Create a clean, simple dataframe for your submission
submission_file <- data.frame(
  Record_ID = 1:length(test_predictions), 
  Predicted_PH = test_predictions
)

# Export the file 
save_path <- "C:/Users/Uzma/CUNY-SPS-Assignments/Data_624/Data_624_Team_Project_2/ABC_Beverage_Final_Predictions.csv"
write.csv(submission_file, save_path, row.names = FALSE)

cat("\n--- File Saved! Now generating visualizations... ---\n")

# A Simple Histogram of your Predictions
hist_plot <- ggplot(submission_file, aes(x = Predicted_PH)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 20) +
  theme_minimal() +
  labs(title = "What the Factory will produce today (Predicted PH)",
       x = "Predicted PH Level",
       y = "Number of Batches")

print(hist_plot)

# The Sanity Check Overlap Plot
train_ph <- data.frame(PH_Value = train_set_final$PH, Data_Type = "1. Historical (Training)")
test_ph <- data.frame(PH_Value = submission_file$Predicted_PH, Data_Type = "2. Predicted (Test)")
combined_ph <- rbind(train_ph, test_ph)

overlap_plot <- ggplot(combined_ph, aes(x = PH_Value, fill = Data_Type)) +
  geom_density(alpha = 0.6) + 
  theme_minimal() +
  scale_fill_manual(values = c("1. Historical (Training)" = "darkgray", 
                               "2. Predicted (Test)" = "darkorange")) +
  labs(title = "Sanity Check: Do the predictions look normal?",
       x = "PH Value",
       y = "Density",
       fill = "Data Source")

print(overlap_plot)
```

**Interpretation***

Chart 1: The Histogram (The Blue Bars)
What it shows: This is your model telling exactly what to expect from the factory today.

The Translation: The vast majority of the beverages produced will have a pH hovering right around 8.5. I can also clearly see a second, smaller cluster over near 8.7. Because we know from our earlier variable importance chart that Brand.CodeC is a important, this second spike is almost certainly making model adjusting the pH perfectly whenever Brand C is running on the line!

Chart 2: The Sanity Check (The Overlap)
What it shows: The gray mountain is the factory's actual history. The orange peak is model's predictions for the new, unseen data.

The Translation: The orange peak fits beautifully inside the footprint of the gray mountain. This proves model is perfectly sane. It didn't hallucinate impossible pH values like 4.0 or 12.0. It respected the physical limits of the machines. The reason the orange peaks are a little sharper is simply because Random Forest is highly confident in its predictions and is locking in on those specific machine set-points.

