---
title: "Project_2_Data_624"
author: "Farhod Ibragimov, Alexander Simon, Jackie Yee, Mubashira Qari"
date: "2026-04-18"
output:
  html_document:
    toc: true
    toc_float: true
  word_document:
    toc: true
---

**As the data science team at ABC Beverage, we are tasked to provide analysis of the company's manufacturing process, the predictive factors and our predictive model of PH to meet compliance of the new regulations.**

```{r message=FALSE, warning=FALSE, include=FALSE}

# Load libraries
library(tidyverse)
library(mice)      # For advanced imputation
library(caret)     # For data splitting and pre-processing
library(rcompanion)
library(ggplot2)
library(UpSetR) # Upset plot
library(corrplot) # corelation plot
library(ggplot2)
library(tidyr)
library(patchwork)
library(kableExtra)
library(naniar)
library(psych)
library(dplyr)
library(rstatix)
library(mice)

# Load the datasets
train_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_624_Team_Project_2/refs/heads/main/StudentData.csv", stringsAsFactors = FALSE)

test_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_624_Team_Project_2/refs/heads/main/StudentEvaluation.csv", stringsAsFactors = FALSE)

```


# Exploratory Data Analysis

## Summary Statistics
**From the summary statistic, we can see that there is 1 categorical variable Brand.Code, our target variable PH, and 31 predictor variables. Additionally, we can see that many of these predictor variables and Brand.Code contain NAs. Notably, in Brand.Code, approximately 48% were categorized into B making the most frequent brand code.**

```{r echo=FALSE}
cat_summary <- train_data %>%
  count(Brand.Code) %>%
  mutate(
    Percentage = round(n / sum(n) * 100, 2)
  ) %>%
  rename(
    "Brand Code" = Brand.Code,
    "Frequency" = n,
    "Percent" = Percentage
  )

cat_summary %>%
  kable(caption = "Descriptive Stats Brand.Code") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
numeric_vars <- train_data %>% select(where(is.numeric))

num_summary <- describe(numeric_vars) %>%
  rownames_to_column("Variable") %>%
  mutate(NA_Count = sapply(numeric_vars, function(x) sum(is.na(x)))) %>%
  select(Variable, n,  mean, sd, median, min, max, NA_Count) %>%
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  rename(
    "Count" = n,
    "Mean" = mean,
    "Std Dev" = sd,
    "Median" = median,
    "Min" = min,
    "Max" = max,
    "NA's" = NA_Count
  )

num_summary %>%
  kable(caption = "Descriptive Stats Numeric Predictors") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

#summary(train_data)


```

## Duplicated Rows Check and NAs check

**We check for any duplicated rows in the train data through the duplicated function. This is to inform us if any duplicated data needs to be handled with appropriately. From the output, we can see that there is no duplicated data. So, next we can begin to check for the percentage of missing data in our dataset, so that we can begin uncovering any patterns related to the missingness.**

```{r}
sum(duplicated(train_data))
```

**From the summary statistic, we can see that there contains many NAs throughout the variables. To check if there is a pattern to these NAs, we can begin by using the [vis_miss() function from the naniar library](https://www.r-bloggers.com/2025/08/handling-missing-data-in-r-a-comprehensive-guide/) to visualize any patterns or relationships that may need to be preserved. In the plot below, we observed that around 1% of the data contains NAs or missing data. 8% of the missing data being from MFR. Other predictors have missingness < 2.5%. Before we can begin to outline with imputation technique to ultilize in dealing with the missing data, we need to look at which predictors correlate to PH through visual analysis.**

```{r echo=FALSE}
vis_miss(train_data) + 
  coord_flip() 
```

```{r eval=FALSE, include=FALSE}
#na_prop <- train_data %>%
#  summarise(across(everything(), ~mean(is.na(.)))) %>%
#  pivot_longer(everything(), names_to = "variable", values_to = "na_proportion")

#ggplot(na_prop, aes(x = reorder(variable, na_proportion), y = na_proportion)) +
#  geom_bar(stat = "identity", aes(fill = na_proportion), show.legend = FALSE) +
#  scale_fill_gradient(low = "lightblue", high = "darkred") +
#  geom_text(aes(label = scales::percent(na_proportion, accuracy = 0.1)), 
#            hjust = -0.1, size = 3) +
#  coord_flip() +
#  theme_minimal(base_size = 12) +
#  theme(panel.grid.major.y = element_blank(),
#        panel.grid.minor = element_blank(),
#        axis.text = element_text(color = "black")) +
#  labs(title = "% of Missing Values by Variable",
#       x = "", 
#       y = "") +
#  scale_y_continuous(labels = scales::percent, limits = c(0, max(na_prop$na_proportion) * 1.1))
```


## Analyzing the Distribution of Predictors

**In the histogram below, our target variable PH is considerably normal. In addition, we can see that there are many ranges of distributions in our histogram of the numeric predictors as some were normal, some were bi-modal and some were skewed. Because of so, we should consider standardization and normalization of predictors in our model building process.**

```{r echo=FALSE, message=FALSE, warning=FALSE}
#histogram
train_data %>%
  select(where(is.numeric)) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "#b19cd9", alpha = 0.7) +
  facet_wrap(~name, scales = "free", ncol = 4) +
  theme_minimal()
```

**Additionally, we check for outliers in the predictors through the box plot below.** 

```{r echo=FALSE, message=FALSE, warning=FALSE}
all_numeric <- train_data %>%
  select(where(is.numeric), -PH) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value")

var_list <- unique(all_numeric$variable)
plots <- list()

for(i in seq_along(var_list)) {
  plots[[i]] <- ggplot(all_numeric %>% filter(variable == var_list[i]), 
                       aes(x = variable, y = value)) +
    geom_boxplot(fill = "#b19cd9", color = "#483d8b", alpha = 0.7,
                 outlier.color = "darkred", outlier.size = 0.8) +
    labs(title = var_list[i], x = "", y = "") +
    theme_minimal(base_size = 8) +
    theme(plot.title = element_text(size = 9, hjust = 0.5),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
}

wrap_plots(plots, ncol = 5)
```

**It is notable that Fill.Ounces, PC.Volume, PSC, Carb.Pressure1, Fill.Pressure, Temperature, MFR, Oxygen.Filler, Air.Pressurer, Filler.Speed contain a large number of outliers mostly skewed, so we will check if these variables have real impact on the target variable PH.**

```{r echo=FALSE, message=FALSE, warning=FALSE}
all_numeric <- train_data %>%
  select(where(is.numeric)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value")

#predictors with high num of outliers
selected_vars <- c("Fill.Ounces", "PC.Volume", "PSC", "Carb.Pressure1", 
                   "Fill.Pressure", "Temperature", "MFR", "Oxygen.Filler", "Air.Pressurer", "Filler.Speed")

plots <- list()

for(i in seq_along(selected_vars)) {
  var <- selected_vars[i]
  plot_data <- all_numeric %>% filter(variable == var)
  
  plots[[i]] <- ggplot(plot_data, aes(x = variable, y = value)) +
    geom_boxplot(fill = "#b19cd9", color = "#483d8b", alpha = 0.7,
                 outlier.color = "darkred", outlier.size = 0.8) +
    labs(title = var, x = "", y = "") +
    theme_minimal(base_size = 8) +
    theme(plot.title = element_text(size = 9, hjust = 0.5),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
}

wrap_plots(plots, ncol = 3)
```

**Upon closer inspection, the predictor variables from the scatter plot below reveals many things in relation to the PH, but first, let's take a look at the predictors Temperature, Air.Pressurer and Filler.Speed.**

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
for(i in seq_along(selected_vars)) {
  var <- selected_vars[i]
  
  plots[[i]] <- ggplot(train_data, aes(x = .data[[var]], y = PH)) +
    geom_point(alpha = 0.4, size = 1) +
    geom_smooth(method = "lm", se = FALSE, color = "darkred") +
    theme_minimal(base_size = 8) +
    labs(title = var, x = "", y = "PH") +
    theme(plot.title = element_text(size = 9, hjust = 0.5))
}

wrap_plots(plots, ncol = 3)
```

**In the individual scatter plot of Temperature, Air.Pressurer and Filler.Speed, they show that they act more like a factor/category than numeric. For these predictors, it could possibly be controlled by a discrete machine or factory setting. For Filler.Speed, the outliers scattered can be presumed to be from the machine alternating between 1000 and 4000. Since these predictors form vertical bands, it is suggested that they don't have high correlation to the PH. However, we can create a correlation table between these predictors and our target variable to verify.**

```{r echo=FALSE}
plot(train_data$Temperature, train_data$PH,
     main = "Temperature vs PH", 
     xlab = "Temperature", ylab = "PH")
plot(train_data$Air.Pressurer, train_data$PH, 
     main = "Air.Pressurer vs PH", 
     xlab = "Air.Pressurer", ylab = "PH")
plot(train_data$Filler.Speed, train_data$PH, 
     main = "Filler.Speed vs PH", 
     xlab = "Filler.Speed", ylab = "PH")
```

**By creating a correlation table, we can see that Air.Pressure and Filler.Speed contain p values > 0.05, signifying no meaningful linear relationship with PH. Temperature has a weak negative correlation to PH with (r = -0.158). While the correlation table shows no meaningful linear relationship, it is not recommended to remove these predictors based on p-values alone, as it disregards non-linear patterns. Additionally, predictors can have meaningful interactions with other predictors to predict PH well. This correlation table is just to test the linear relationships the predictors have on PH and to inform the outlier detection process.**

```{r echo=FALSE}
vars <- c("Temperature", "Air.Pressurer", "Filler.Speed")

cor_table <- data.frame()

for(var in vars) {
  test <- cor.test(train_data[[var]], train_data$PH, use = "complete.obs")
  cor_table <- rbind(cor_table, data.frame(
    Variable = var,
    Correlation = round(test$estimate, 3),
    P_Value = test$p.value,
    Significant = ifelse(test$p.value < 0.05, "Yes", "No")
  ))
}

cor_table$P_Value <- format(cor_table$P_Value, scientific = TRUE)
print(cor_table)
```

**Afterwards, we can begin to do more analysis of the predictors left: Temperature, Fill.Ounces, PC.Volume, PSC, Carb.Pressure1, Oxygen.Filler, MFR and Fill.Pressure. We will use the [IQR rule](https://www.r-bloggers.com/2025/12/outliers-in-data-analysis-detecting-extreme-values-before-modeling-in-r-with-istanbul-airbnb-data/) for detecting outliers outside the IQR range (Q1 - 1.5 * IQR AND Q3 + 1.5 * IQR).**

**From the graph, we see:**
- `Fill.Ounces`, `PC.Volume`, `PSC`, `Carb.Pressure1` and `Fill.Pressure` have outliers that aren't extremely far from the IQR.
- `Oxygen.Filler`, `MFR` and `Temperature` suggests extreme outliers.

**From our previous observations, we noted that Temperature acts more like a factor/category than numeric predictor.**

```{r echo=FALSE, message=FALSE, warning=FALSE}
vars <- c("Fill.Ounces", "PC.Volume", "PSC", "Carb.Pressure1", "Oxygen.Filler", "MFR", "Fill.Pressure", "Temperature")

plots <- list()

for(var in vars) {
  #iqr
  Q1 <- quantile(train_data[[var]], 0.25, na.rm = TRUE)
  Q3 <- quantile(train_data[[var]], 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower <- Q1 - 1.5 * IQR
  upper <- Q3 + 1.5 * IQR
  
  #flagging the outliers
  plot_data <- train_data %>%
    mutate(is_outlier = train_data[[var]] < lower | train_data[[var]] > upper,
           outlier_color = ifelse(is_outlier, "Outlier", "Normal"))
  
  p <- ggplot(plot_data, aes(x = .data[[var]], y = PH, color = outlier_color)) +
    geom_point(alpha = 0.6, size = 1) +
    geom_smooth(method = "lm", color = "#000000", se = FALSE, size = 0.8) + 
    scale_color_manual(values = c("Normal" = "#b19cd9", "Outlier" = "darkred")) +
    theme_minimal(base_size = 8) +
    theme(legend.position = "none",
          plot.title = element_text(size = 9, hjust = 0.5)) +
    labs(title = var, x = "", y = "PH")
  
  plots[[var]] <- p
}

wrap_plots(plots, ncol = 4)
```

**The above scatterplot has overplotting issues due to the data points clustering in a cloud-like formation, so we decided on randomly sampling 500 rows of the dataset to reduce this issue. From the randomly sampled scatter plot of the predictors, we can see that:**
- `MFR` and `Fill.Pressure` exhibits factor/category likeness, rather than numeric likeness. Distribution across the PH regression line suggestions no to weak correlation.
- `Fill.Ounces`, `PC.Volume`, `PSC`, `Carb.Pressure1` and `Oxygen.Filler` points on the scatterplot exhibit weak/no relationship to the regression line of PH.


```{r echo=FALSE, message=FALSE, warning=FALSE}
set.seed(5555)
sample_data <- train_data %>% sample_n(500)

plots <- list()

for(var in vars) {
  Q1 <- quantile(train_data[[var]], 0.25, na.rm = TRUE)
  Q3 <- quantile(train_data[[var]], 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower <- Q1 - 1.5 * IQR
  upper <- Q3 + 1.5 * IQR
  
  plot_data <- sample_data %>%
    mutate(is_outlier = sample_data[[var]] < lower | sample_data[[var]] > upper,
           outlier_color = ifelse(is_outlier, "Outlier", "Normal"))
  
  p <- ggplot(plot_data, aes(x = .data[[var]], y = PH, color = outlier_color)) +
    geom_point(alpha = 0.6, size = 1) +
    geom_smooth(method = "lm", color = "darkred", se = FALSE, size = 1) +
    scale_color_manual(values = c("Normal" = "#b19cd9", "Outlier" = "darkred")) +
    theme_minimal(base_size = 8) +
    theme(legend.position = "none",
          plot.title = element_text(size = 9, hjust = 0.5)) +
    labs(title = var, x = "", y = "PH")
  
  plots[[var]] <- p
}

wrap_plots(plots, ncol = 4)
```

**We can additionally check MFR and Fill.Pressure for their correlation to PH as they exhibit vertical clustering. From the table, we can see that MFR has a p values > 0.05, signifying no meaningful linear relationship with PH. While Fill.Pressure's p-value is < 0.05 and signifies linear correlation to PH.**

```{r echo=FALSE}
vars <- c("MFR", "Fill.Pressure")

cor_table <- data.frame()

for(var in vars) {
  test <- cor.test(train_data[[var]], train_data$PH, use = "complete.obs")
  cor_table <- rbind(cor_table, data.frame(
    Variable = var,
    Correlation = round(test$estimate, 3),
    P_Value = test$p.value,
    Significant = ifelse(test$p.value < 0.05, "Yes", "No")
  ))
}

cor_table$P_Value <- format(cor_table$P_Value, scientific = TRUE)
print(cor_table)
```

**From analyzing the predictors with seemingly large number of outliers or extreme outliers, it should be noted that none of the tested predictors have a notable significance to PH. Additionally, predictors such as MFR, Fill.Pressure, Temperature, Air.Pressure and Filler.Speed exhibits factor/category likeness, rather than numeric likeness and their range outliers most likely stem from the machine or factory setting.**

## Analyzing Brand.Code

**Afterwards, we can check if the categorical variable Brand.Code has any significant predictive effects on the target variable PH. This is to inform us whether we should keep our categorical variable or not. Through the boxplot, we can see that each brand (including NAs) have varying median lines, indicating that they are different types of beverages that potentially come from various manufacturing processes. Interestingly, the NAs Brand.Code distribution ties closely to the distribution of all over brand.codes.**


```{r echo=FALSE}
boxplot(PH ~ Brand.Code, data = train_data, 
        las = 2, xlab = "Brand Code", ylab = "PH", 
        main = "PH Distribution by Brand Code")
```

**Let's confirm that each brand has a statistically different PH distribution from every other brand through a [mann-Whitney U test](https://www.statology.org/mann-whitney-u-test/). As shown below, we can see that each pairwise comparison of Brand.Code has a statistically significant pH value (all p-values < 0.05 after [Bonferroni correction](https://www.statology.org/bonferroni-correction/)), indicting that the brands represent different beverage types or formulations, not only a minor change/variation.**

```{r echo=FALSE, message=FALSE, warning=FALSE}
#mann-Whitney U test pairwise comparisons, labeled brand.code nas as missing
ph_by_brand <- train_data %>%
  mutate(Brand.Code = forcats::fct_na_value_to_level(Brand.Code, "Missing"))

#https://www.statology.org/bonferroni-correction/
#pairwise Mann-Whitney U test with bonferroni
#in Bonferroni correction each raw p-value is multiplied by ( m ) (compare each p-value to (\alpha/m)).
#ensures strong control of the family-wise error rate.
pairwise_results <- pairwise.wilcox.test(
  ph_by_brand$PH, 
  ph_by_brand$Brand.Code,
  p.adjust.method = "bonferroni",
  paired = FALSE
)

print(pairwise_results)
```

**From an ANOVA test of the Brand.Code to see the effects of Brand.Code on PH, we can see that p < 2e-16, signifying it as highly significant. Additionally, Brand.Code explains 11.7% of the variance in PH. Because of this, we decided to keep Brand.Code.**

```{r echo=FALSE}
aov_result <- aov(PH ~ Brand.Code, data = train_data)
summary(aov_result)

ss <- summary(aov_result)[[1]]$"Sum Sq"
r_sq <- ss[1] / sum(ss)
#variance by brand.code
round(r_sq * 100, 2)
```

## Analyzing Variable Correlations

**Before we begin to check the correlation of predictors, it is important to exclude PH from the training data so that it doesn't affect any of the predictor correlations. After doing so, we can start analyzing the relationships of the variables. Through the correlation plot below, we can see that variables such as Hy.Pressure3, Filler.Level, Filler.Speed, Density, Balling, Alch.Rel and Carb.Rel have correlations > 85%. Before we remove the variables, however, we can use the function findCorrelation() to match our correlation plot.**

```{r echo=FALSE}
#excludingPH from numeric data
numeric_data <- train_data %>% 
  select(where(is.numeric), -PH)

cor_matrix <- cor(numeric_data, use = "complete.obs")

cor_matrix_filtered <- cor_matrix
cor_matrix_filtered[abs(cor_matrix_filtered) < 0.85] <- NA

corrplot(cor_matrix_filtered, 
         method = "color", 
         type = "upper", 
         tl.cex = 0.7,
         na.label = " ",
         col = colorRampPalette(c("darkred", "white", "#483d8b"))(200),
         title = "Correlations |r| > 0.85",
         mar = c(0,0,2,0))
```

**From the findCorrelation function, we can see that they match the analysis of the correlation plot, with the exception of Carb.Rel. This is likely due to carb.rel having an average correlation that is less than 85% with the other predictors. As such, it will not be a candidate for removal. However, the rest of the predictors: "Balling", "Hyd.Pressure3", "Alch.Rel", "Balling.Lvl", "Density", "Filler.Level" and "Filler.Speed" is recommended.**

```{r echo=FALSE}
high_corr <- findCorrelation(cor_matrix, cutoff = 0.85, names = TRUE, exact = TRUE)
high_corr 

high_cor_pairs <- which(abs(cor_matrix) > 0.85 & abs(cor_matrix) < 1, arr.ind = TRUE)
high_cor_pairs <- data.frame(
  Var1 = rownames(cor_matrix)[high_cor_pairs[,1]],
  Var2 = colnames(cor_matrix)[high_cor_pairs[,2]],
  Corr = cor_matrix[high_cor_pairs]
) %>% filter(Var1 != Var2) %>% arrange(desc(abs(Corr)))

#(high_cor_pairs)
```

## Reccomendations from EDA

- Predictors: `Balling`, `Hyd.Pressure3`, `Alch.Rel`, `Balling.Lvl`, `Density`,`Filler.Level`, and `Filler.Speed` are suggested to be removed due to no significant correlation to other predictors.
- Consider robust methods or tree-based models over standard linear regression when dealing with the predictor outliers. Several predictors (`MFR`, `Fill.Pressure`, `Temperature`, `Air.Pressure`, `Filler.Speed`) show non-normal distributions and non/weak linear correlations with PH. In addition to exhibiting patterns as that of an ordinal categorical variable. This approach could be less sensitive to outliers and can capture non-linear patterns without requiring transformations.
- After removal of suggested predictors, it is recommended to use multiple imputation (such as missForest) on NAs. As it is flexible to our sample size of predictors. Since we also will be keeping our categorical variable, `Brand.Code`, missForest allows for working with various kinds of continuous and categorical variables, creating a single completed data set by iteratively predicting missing values.
- Consider standardization and normalization of predictors in the model building process due to as predictors exhibit bimodal and skewed distribution.

**Note: Brand.Code most likely represents different beverage types or formulations, not only a minor change/variation. Additionally, beverages for PH are not typically above 9. Any beverage above that PH can be considered alkaline and are unusual, but they do exist.**


=========================================================================
**Data Preparation: Imputation, transformations, Encoding and Filtering**
=========================================================================


```{r}
library(skimr)

skim(train_data)
```


```{r}

# View the structure and summary of the data
#str(train_data)
#summary(train_data)
```

**Target Variable (PH) Validation**

First we examine the PH column in the training set and remove records that do not have an associated PH value.
If we include rows in our training set where the PH value is missing (NA), the mathematical algorithms will have no answer key to learn from for those specific rows. 

While we can use statistical methods to guess (impute) missing predictor values like temperature, we should never guess the target value in the training data, since this introduces artificial bias into our model. 

Removing these 4 records ensures our model is built only on confirmed, historical production facts.

```{r}
# Count missing values in the target variable
missing_ph <- sum(is.na(train_data$PH))
print(paste("Missing PH values in training set:", missing_ph))

# Remove rows where PH is NA
filtered_data <- train_data |>
  filter(!is.na(PH))

head(filtered_data)
```

**Column Name Standardization**

```{r}
# Replace all periods with underscores in the column names
colnames(filtered_data) <- gsub("\\.", "_", colnames(filtered_data))
colnames(test_data) <- gsub("\\.", "_", colnames(test_data))

# Quick check to see the new names
names(filtered_data)
```


**Analysis of Dataset**

```{r message=FALSE, warning=FALSE}
# Checking the total number of missing values per column
missing_summary <- colSums(is.na(filtered_data))
print(missing_summary[missing_summary > 0]) # Show only columns with NAs

# Visualizing the missing patterns
# install.packages("VIM")
library(VIM)
aggr(filtered_data, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, 
     labels=names(train_data), cex.axis=.7, gap=3, 
     ylab=c("Histogram of missing data","Pattern"))
```


```{r}
head(filtered_data)
```

In Chapter 3 of Applied Predictive Modeling, the nearZeroVar function uses a strict two-part mathematical test to decide if a predictor is useless (NZV). A predictor is only marked TRUE if it satisfies both of these conditions:

The Frequency Ratio is > 19: This means the most common value appears at least 19 times more often than the second most common value.

The Percent Unique is < 10%: This means the number of unique values in the column is less than 10% of the total number of rows.

**Identify Near-Zero Variance (NZV) Predictors**

```{r}

# 'saveMetrics = TRUE' provides the detailed math behind the decision
nzv_analysis <- nearZeroVar(filtered_data, saveMetrics = TRUE)

nzv_analysis
```
When we look at the raw data for Hyd Pressure1, we can see exactly why it matched both conditions:

Frequency Ratio (~31.1): The value 0.0 appears 840 times. 
Since 840 / 27 approx. 31.1, it far exceeds the imbalance threshold of 19 and this is below 10%. 
This tells the algorithm that this variable is almost a constant zero.

Consequence of not removing: removing Hyd Pressure1 because if you perform Resampling like Cross-Validation, there is a high statistical chance that in one of folds, that variable will become a constant 0 for every single row. If that happens, many models will crash because they cannot calculate a slope or weight for a column that never changes.


```{r}
# Identify the indices of the predictors to remove

nzv_indices <- nearZeroVar(filtered_data)

# Remove these predictors from the TRAIN and TEST sets
# Saving into NEW variables as requested
train_data_cleaned <- filtered_data[, -nzv_indices]
test_data_cleaned <- test_data[, -nzv_indices]

# Verify the change
# Check how many columns were removed
cat("Original number of columns:", ncol(filtered_data), "\n")
cat("New number of columns:", ncol(train_data_cleaned), "\n")

# View which column was removed
colnames(filtered_data)[nzv_indices]
```


```{r}
# Identifying which column index corresponds to the NZV predictor 
# from our previous nzv_analysis object
nzv_cols <- which(nzv_analysis$nzv == TRUE)

# Removing the columns from train and test datasets

train_data_final <- filtered_data[, -nzv_cols]
test_data_final <- test_data[, -nzv_cols]

# Verifying that the shapes match
dim(train_data_final)
dim(test_data_final)

```

**Transformations to Resolve Skewness**

Many predictive models (especially linear models and Neural Networks) perform much better when the predictors have a symmetric, normal distribution.


**Checking Skewness**

```{r}
# Load required libraries
library(tidyverse) # Contains dplyr, tidyr, and ggplot2
library(e1071)     # Contains the skewness() function

# Starting the pipeline with your cleaned training data
train_data_final |>
  
  # Keeping only the numeric columns
  select(where(is.numeric)) |>
  
  # Calculating skewness for every remaining column
  summarise(across(everything(), ~skewness(.x, na.rm = TRUE))) |>
  
  # Reshaping the data from wide (one row) to long (two columns: Predictor, Skewness)
  pivot_longer(cols = everything(), names_to = "Predictor", values_to = "Skewness") |>
  
  # Adding the flag to identify which variables are highly skewed
  mutate(Highly_Skewed = ifelse(abs(Skewness) > 1, "High Skew", "Acceptable Skew")) |>
  
  # Pipe the formatted data directly into ggplot
  ggplot(aes(x = reorder(Predictor, Skewness), y = Skewness, fill = Highly_Skewed)) +
  geom_col(color = "black", linewidth = 0.2) +
  coord_flip() + 
  scale_fill_manual(values = c("Acceptable Skew" = "skyblue", "High Skew" = "salmon")) +
  
  # Add the safe threshold lines
  geom_hline(yintercept = c(-1, 1), linetype = "dashed", color = "red", linewidth = 1) +
  
  labs(
    title = "Skewness of All Numeric Predictors",
    subtitle = "Variables extending past the red dashed lines need Box-Cox transformation",
    x = "Predictors",
    y = "Skewness Value"
  ) +
  theme_minimal()
```

The chart identifies which numeric predictors in dataset have a high skewness extending beyond the safe +/- 1 threshold (red lines), highlighting exactly which variables using Yeo-Johnson instead of Box-Cox to fix skewed distributions.

**Handle Skewness in Data**

```{r}
library(caret)

# Create a pre-processing plan using Yeo-Johnson instead of Box-Cox
# This will handle the zeros in variables like Filler_Speed and Mnf_Flow
yj_plan <- preProcess(
  train_data_final[, !(names(train_data_final) %in% c("PH", "Brand_Code"))], 
  method = "YeoJohnson"
)

# Apply the plan to fix the skewness
train_data_unskewed <- predict(yj_plan, train_data_final)
test_data_unskewed <- predict(yj_plan, test_data_final)

# 3. Check the skewness again to verify it worked this time!
library(tidyverse)
library(e1071)

train_data_unskewed |>
  select(where(is.numeric)) |>
  summarise(across(everything(), ~skewness(.x, na.rm = TRUE))) |>
  pivot_longer(cols = everything(), names_to = "Predictor", values_to = "New_Skewness") |>
  arrange(desc(abs(New_Skewness))) |>
  head(10)
```

**Verify Skewness**

```{r}
# Load required libraries
library(tidyverse)
library(e1071)

# Check the skewness of the newly unskewed dataset
train_data_unskewed |>
  
  # Keep only the numeric columns
  select(where(is.numeric)) |>
  
  # Calculate skewness for every remaining column
  summarise(across(everything(), ~skewness(.x, na.rm = TRUE))) |>
  
  # Reshape into a clean two-column table
  pivot_longer(cols = everything(), names_to = "Predictor", values_to = "New_Skewness") |>
  
  # Sort by the highest absolute skewness to see the "worst" ones first
  arrange(desc(abs(New_Skewness))) |>
  
  # Print the top 10 results
  head(10)
```

```{r}
# Load required libraries
library(tidyverse)
library(e1071)

# Start the pipeline with the dataset AFTER the Yeo-Johnson transformation
train_data_unskewed |>
  
  # Keep only the numeric columns
  select(where(is.numeric)) |>
  
  # Calculate the NEW skewness for every remaining column
  summarise(across(everything(), ~skewness(.x, na.rm = TRUE))) |>
  
  # Reshape the data for plotting
  pivot_longer(cols = everything(), names_to = "Predictor", values_to = "Skewness") |>
  
  # Add the flag to identify which variables are STILL highly skewed
  mutate(Highly_Skewed = ifelse(abs(Skewness) > 1, "High Skew (Still)", "Acceptable Skew")) |>
  
  # Pipe into ggplot for the visualization
  ggplot(aes(x = reorder(Predictor, Skewness), y = Skewness, fill = Highly_Skewed)) +
  geom_col(color = "black", linewidth = 0.2) +
  coord_flip() + 
  scale_fill_manual(values = c("Acceptable Skew" = "skyblue", "High Skew (Still)" = "salmon")) +
  
  # Add the safe threshold lines
  geom_hline(yintercept = c(-1, 1), linetype = "dashed", color = "red", linewidth = 1) +
  
  labs(
    title = "Skewness AFTER Yeo-Johnson Transformation",
    subtitle = "Variables extending past the red lines",
    x = "Predictors",
    y = "New Skewness Value"
  ) +
  theme_minimal()
```
```{r}
# Load the required library
library(tidyverse)

# Define the updated list of the 6 stubborn variables
# (Ensure the spelling exactly matches your dataset column names)
stubborn_vars <- c("MFR", "Filler_Speed", "Oxygen_Filler", "Temperature", "Air_Pressurer", "PSC_CO2")

# 2. Build the pipeline to plot all six automatically
train_data_final |>
  
  # Keep only these 6 specific columns
  select(all_of(stubborn_vars)) |>
  
  # Reshape the data so ggplot can draw them all at once
  pivot_longer(cols = everything(), names_to = "Predictor", values_to = "Raw_Value") |>
  
  # Pipe directly into the plot
  ggplot(aes(x = Raw_Value)) +
  geom_histogram(bins = 50, fill = "salmon", color = "black", alpha = 0.8) +
  
  # facet_wrap splits the plot into 6 separate panels
  # 'scales = "free"' is critical here so each chart gets its own correct axis
  facet_wrap(~ Predictor, scales = "free") +
  
  labs(
    title = "Raw Distributions of the Top 6 Most Skewed Variables",
    subtitle = "Diagnosing why mathematical transformations failed",
    x = "Raw Values",
    y = "Count of Rows"
  ) +
  theme_minimal()
```
**Visualizing Outliers for Stubborn Variables**

```{r}
# Load the required library
library(tidyverse)

#  Define the list of stubborn variables
stubborn_vars <- c("MFR", "Filler_Speed", "Oxygen_Filler", "Temperature", "Air_Pressurer", "PSC_CO2")

# Build the pipeline for Boxplots
train_data_final |>
  # Select the variables
  select(all_of(stubborn_vars)) |>
  
  # Reshape for ggplot
  pivot_longer(cols = everything(), names_to = "Predictor", values_to = "Raw_Value") |>
  
  # Create the plot
  ggplot(aes(x = Predictor, y = Raw_Value, fill = Predictor)) +
  # geom_boxplot is the gold standard for outlier detection
  geom_boxplot(outlier.color = "red", outlier.shape = 16, outlier.size = 2) +
  
  # Separate into individual panels for clarity
  facet_wrap(~ Predictor, scales = "free") +
  
  labs(
    title = "Outlier Detection for Stubborn Predictors",
    subtitle = "Red dots represent potential outliers (values beyond 1.5 * IQR)",
    x = "",
    y = "Raw Value Range"
  ) +
  theme_minimal() +
  theme(legend.position = "none") # Hide legend since facets are labeled
```
Interpretation:

Zero-Inflated Predictors (MFR, Filler_Speed, Air_Pressurer): The data is concentrated at a single value (likely zero), causing the boxplots to appear flattened. Since these lack the necessary spread for mathematical transformations to work, they should be evaluated as Near-Zero Variance (NZV) predictors and either removed or converted into binary "On/Off" indicators.

Heavy-Tailed Predictors (Oxygen_Filler, Temperature): These show a string of pearls pattern with many outliers extending upward. Because these extreme values pull the skewness calculation heavily to the right, handling involves Winsorizing (capping values at a specific percentile) or using a Spatial Sign Transformation to minimize outlier influence.

Discontinuous Predictors (PSC_CO2): This variable shows a distinct set of outliers far above the main data cluster. Standard transformations fail to bridge this gap, so the best approach is to either treat the variable as a categorical factor or use tree-based models that are inherently robust to such gaps.


```{r}
library(tidyverse)

# Define the six stubborn variables
stubborn_vars <- c("MFR", "Filler_Speed", "Oxygen_Filler", "Temperature", "Air_Pressurer", "PSC_CO2")

# Calculate outliers per column
outlier_report <- train_data_final |>
  select(all_of(stubborn_vars)) |>
  pivot_longer(cols = everything(), names_to = "Predictor", values_to = "Value") |>
  group_by(Predictor) |>
  summarise(
    Q1 = quantile(Value, 0.25, na.rm = TRUE),
    Q3 = quantile(Value, 0.75, na.rm = TRUE),
    IQR = Q3 - Q1,
    Lower_Bound = Q1 - 1.5 * IQR,
    Upper_Bound = Q3 + 1.5 * IQR,
    Outlier_Count = sum(Value < Lower_Bound | Value > Upper_Bound, na.rm = TRUE),
    Percent_Outliers = round((Outlier_Count / n()) * 100, 2)
  )

print("--- Outlier Count Per Predictor ---")
print(outlier_report)

# Calculate the 'Total Impact' (Rows with at least one outlier)
rows_with_outliers <- train_data_final |>
  mutate(has_outlier = rowSums(across(all_of(stubborn_vars), function(x) {
    q <- quantile(x, c(0.25, 0.75), na.rm = TRUE)
    iqr <- q[2] - q[1]
    x < (q[1] - 1.5 * iqr) | x > (q[2] + 1.5 * iqr)
  }), na.rm = TRUE) > 0) |>
  summarise(
    Total_Impacted_Rows = sum(has_outlier),
    Total_Rows_In_Dataset = n(),
    Percent_Data_Loss = round((Total_Impacted_Rows / n()) * 100, 2)
  )

print("--- Total Impact on Dataset ---")
print(rows_with_outliers)
```
A 33.66% data loss is far too high; deleting one-third of the observations would significantly weaken the model's ability to learn and generalize.

Interpretation of Stubborn Predictors

Filler_Speed (17.06% Outliers): This high percentage confirms a multi-modal distribution. The "outliers" are likely the machine's actual operating speeds, while the majority of the data represents the machine being idle or in a secondary mode.

MFR (9.27% Outliers): This follows a zero-inflated pattern. The boxplot is squashed because the process stays at a fixed baseline most of the time, making any actual production variance appear as a mathematical outlier.

Air_Pressurer (8.57% Outliers): This variable has a highly concentrated core. Because the normal operating range is extremely narrow, even standard, valid fluctuations are being flagged as red dots.

Oxygen_Filler (7.25% Outliers): This represents a heavy-tailed distribution. There is a string of points extending toward higher values, which pulls the mean away from the median and creates persistent skewness.

Temperature (4.75% Outliers): Similar to Oxygen_Filler, this has a right-skewed tail. While the percentage is lower, these extreme heat spikes can disproportionately influence linear model coefficients.

PSC_CO2 (3.00% Outliers): This shows a discontinuous pattern. There is a small group of rows where CO2 levels are significantly higher than the rest of the production run, creating a gap in the data.

**Strategy for Handling Outliers**

Since losing 33.66% of the data by deleting rows is unacceptable, the following "Soft Fix" approach from Chapter 3 is recommended:

Winsorizing (Capping): Set the values for all six variables to their 1st and 99th percentiles. This brings the extreme red dots to the edge of the acceptable range, neutralizing their negative impact on the model while keeping every single row of data.

Spatial Sign Transformation: After capping, we apply the Spatial Sign. This projects the data onto a multidimensional sphere, ensuring that any remaining distance from the outliers doesn't overwhelm the model's loss function.

Binary Indicator for Zero-Inflation: For Filler_Speed and MFR, we can consider adding a new column (e.g., Is_Running) to help the model distinguish between "Off" and "Active" states more clearly than a skewed number can.

```{r}
# Define a robust capping function
cap_outliers <- function(x) {
  # Calculate the 1st and 99th percentiles
  qnts <- quantile(x, probs = c(0.01, 0.99), na.rm = TRUE)
  
  # Replace values outside those bounds
  x[x < qnts[1]] <- qnts[1]
  x[x > qnts[2]] <- qnts[2]
  return(x)
}

# Define the list of stubborn variables
stubborn_vars <- c("MFR", "Filler_Speed", "Oxygen_Filler", 
                   "Temperature", "Air_Pressurer", "PSC_CO2")

# Apply the cap to create a new dataset
train_data_capped <- train_data_final |>
  mutate(across(all_of(stubborn_vars), cap_outliers))

# Apply the same logic to the test set 
test_data_capped <- test_data_final |>
  mutate(across(all_of(stubborn_vars), cap_outliers))
```

**Visualizing the Impact of Capping**

```{r}
library(tidyverse)
library(e1071)
library(gridExtra)

# Prepare data for Skewness Comparison
# Helper function to get skewness for the 6 stubborn variables
calc_skew <- function(df, label) {
  df |>
    select(all_of(stubborn_vars)) |>
    summarise(across(everything(), ~skewness(.x, na.rm = TRUE))) |>
    pivot_longer(cols = everything(), names_to = "Predictor", values_to = "Skewness") |>
    mutate(Version = label)
}

# Combine results
skew_compare <- bind_rows(
  calc_skew(train_data_final, "Original (Salmon)"),
  calc_skew(train_data_capped, "Capped (Blue)")
)

# Skewness Bar Chart Comparison
plot_skew <- ggplot(skew_compare, aes(x = reorder(Predictor, abs(Skewness)), y = Skewness, fill = Version)) +
  geom_col(position = "dodge", color = "black", linewidth = 0.2) +
  coord_flip() +
  geom_hline(yintercept = c(-1, 1), linetype = "dashed", color = "red", linewidth = 1) +
  scale_fill_manual(values = c("Original (Salmon)" = "salmon", "Capped (Blue)" = "skyblue")) +
  labs(
    title = "Impact of Capping on Skewness",
    subtitle = "The Blue bars should be significantly shorter than the Salmon bars",
    x = "Predictors", y = "Skewness Value"
  ) +
  theme_minimal()

# Boxplot Comparison (example: Oxygen_Filler)
# This shows how the 'string of pearls' is handled
boxplot_compare <- bind_rows(
  train_data_final |> select(Oxygen_Filler) |> mutate(Version = "Original"),
  train_data_capped |> select(Oxygen_Filler) |> mutate(Version = "Capped")
) |>
  ggplot(aes(x = Version, y = Oxygen_Filler, fill = Version)) +
  geom_boxplot(outlier.color = "red") +
  scale_fill_manual(values = c("Original" = "salmon", "Capped" = "skyblue")) +
  labs(title = "Outlier Compression", subtitle = "Oxygen_Filler Before vs After") +
  theme_minimal() + theme(legend.position = "none")

# Display both
grid.arrange(plot_skew, boxplot_compare, ncol = 1)
```
While Capping (Winsorizing) squashed the most extreme values, the Spatial Sign is the specialized tool suggested in the Kuhn & Johnson chapter 3 of Applied Predictive Modeling, designed to neutralize the remaining influence of those red dots (outliers).

Spatial Sign Transformation pulls every outlier toward the center of the data mass. It doesn't delete them; it just ensures they don't have enough weight to pull your model's predictions off-track.

For the Spatial Sign to work correctly, the data must be Centered and Scaled first otherwise the sphere will be skewed and the transformation will actually make your data worse.

**Applying the Spatial Sign Transformation, Scaling and Centering**

```{r}
library(caret)

# We must include 'center' and 'scale' for 'spatialSign' to work
spatial_plan <- preProcess(
  train_data_capped |> select(-PH, -Brand_Code), 
  method = c("center", "scale", "spatialSign")
)

# Apply the transformation to both datasets
train_data_spatial <- predict(spatial_plan, train_data_capped)
test_data_spatial <- predict(spatial_plan, test_data_capped)

# Check the range of a stubborn variable
# All values should now be between -1 and 1
summary(train_data_spatial$Oxygen_Filler)

```

**Visual Proof of Outlier Neutralization**

```{r}
library(tidyverse)

# 1. Gather the three versions of the data into one long format dataframe
comparison_data <- bind_rows(
  train_data_final   |> select(all_of(stubborn_vars)) |> mutate(Stage = "1. Original"),
  train_data_capped  |> select(all_of(stubborn_vars)) |> mutate(Stage = "2. Capped"),
  train_data_spatial |> select(all_of(stubborn_vars)) |> mutate(Stage = "3. Spatial Sign")
) |>
  pivot_longer(cols = -Stage, names_to = "Predictor", values_to = "Value") |>
  mutate(Stage = factor(Stage, levels = c("1. Original", "2. Capped", "3. Spatial Sign")))

# 2. Create the comparison visualization
ggplot(comparison_data, aes(x = Stage, y = Value, fill = Stage)) +
  geom_boxplot(outlier.color = "red", outlier.shape = 16, outlier.size = 1.2, alpha = 0.7) +
  # Scales = "free_y" is critical because the Spatial Sign scale is much smaller (-1 to 1)
  facet_wrap(~ Predictor, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = c(
    "1. Original" = "salmon", 
    "2. Capped" = "skyblue", 
    "3. Spatial Sign" = "mediumpurple1"
  )) +
  labs(
    title = "The Impact of Outlier Neutralization",
    subtitle = "Observe how the red dots are pulled toward the center in Stage 3",
    x = "",
    y = "Calculated Value (Note the scale changes per stage)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )
```
Stage 1: Original (The Danger Zone): Raw data contains outliers with massive vertical tails. These extreme values exert too much leverage, forcing the model to skew its logic just to accommodate a few rogue points.

Stage 2: Capped (The Initial Taming): Applying Winsorization (99th percentile capping) pulls the extreme into a set boundary. This removes their excessive leverage while keeping the rows in the dataset.

Stage 3: Spatial Sign (The Golden State): The data is projected onto a multidimensional sphere. This centers the variables and brings remaining outliers so they are mathematically close to the mean, ensuring they no longer have the weight to pull the model off track.


Earlier we saw that removing rows with outliers would have cost us 33.66% of your data. 
Missing values create a similar problem in modeling the data.

If 10% of rows have at least one NA, deleting those rows shrinks training data significantly.
By imputing, we keep all 2,567 rows, which gives the model more examples to learn from, making it more robust and accurate.

```{r}
head(train_data_spatial)
```


**Feature Engineering: Binary Flags for Zero-Inflation**

```{r}
# We use the original data to identify 'True Zeros' before imputation fills them.
# This explicitly tells the model when a machine was IDLE.
train_data_spatial <- train_data_spatial |>
  mutate(
    MFR_Active = ifelse(train_data_final$MFR == 0, 0, 1),
    Filler_Active = ifelse(train_data_final$Filler_Speed == 0, 0, 1),
    Air_Pressurer_Active = ifelse(train_data_final$Air_Pressurer == 0, 0, 1)
  )

test_data_spatial <- test_data_spatial |>
  mutate(
    MFR_Active = ifelse(test_data_final$MFR == 0, 0, 1),
    Filler_Active = ifelse(test_data_final$Filler_Speed == 0, 0, 1),
    Air_Pressurer_Active = ifelse(test_data_final$Air_Pressurer == 0, 0, 1)
  )

#test_data_spatial
```

**Dealing with Missing Values (KNN Imputation)**

```{r}
# We fill the 208 gaps in MFR and others based on the 'neighborhood' of similar rows.
# Note: knnImpute also centers and scales, but since we are already spatial-signed,
# it will just focus on filling the holes.
impute_proc <- preProcess(train_data_spatial |> select(-PH, -Brand_Code), method = "knnImpute")

train_data_complete <- predict(impute_proc, train_data_spatial)
test_data_complete  <- predict(impute_proc, test_data_spatial)

# Verify: This should return 0
cat("Total NAs remaining in Training Data:", sum(is.na(train_data_complete)), "\n")
```


**The "Pre-Encoding" Check**

```{r}
# Count total NAs in Brand_Code
cat("Total NULLs in Brand_Code:", sum(is.na(train_data_complete$Brand_Code)), "\n")

# View distribution including the NA count
# 'useNA = "always"' ensures we see the Nulls as their own category
table(train_data_complete$Brand_Code, useNA = "always")
```
The result shows that Brand_Code has 0 literal NULLs, but it does have 120 Hidden Nulls

In R, empty strings ("") aren't technically NA, but for a predictive model, they are just as problematic. We need to handle those blanks before turning the brands into numeric columns.

```{r}
train_data_complete
```

**Resolving Blanks & Generating the Heatmap**

```{r}
# Resolve the 120 blanks in Brand_Code (before they become a dummy column)
# We map the empty strings to Brand 'B' (the Mode)
train_data_complete$Brand_Code[train_data_complete$Brand_Code == ""] <- "B"
test_data_complete$Brand_Code[test_data_complete$Brand_Code == ""]   <- "B"

# Re-run Dummy Encoding on the fixed Brand_Code
dummies <- dummyVars(PH ~ ., data = train_data_complete)
train_data_encoded <- as.data.frame(predict(dummies, newdata = train_data_complete))
train_data_encoded$PH <- train_data_complete$PH

# Calculate the Correlation Matrix (Excluding the target PH)
# We use 'use = "pairwise.complete.obs"' as a safety measure
cor_matrix <- cor(train_data_encoded |> select(-PH), use = "pairwise.complete.obs")

# Handle any potential NAs in the correlation matrix 
# We find columns that have 0 variance (constant columns)
zero_var_cols <- which(apply(train_data_encoded |> select(-PH), 2, sd) == 0)
if(length(zero_var_cols) > 0) {
  train_data_encoded <- train_data_encoded[, -zero_var_cols]
  cor_matrix <- cor(train_data_encoded |> select(-PH))
}

# Generate the Correlation Heatmap
library(corrplot)
corrplot(cor_matrix, 
         method = "color", 
         type = "upper", 
         order = "hclust", # Groups redundant "twin" variables together
         tl.col = "black", 
         tl.srt = 45, 
         tl.cex = 0.5,     # Smaller text to fit all predictors
         title = "Predictor Multicollinearity: Before Filtering",
         mar = c(0,0,1,0))
```
We could map those 120 blanks to a new label called "Unknown", however according to book: Applied Predictive Modeling chapter 3 by Kuhn & Johnson, if the number of missing values is small, simple imputation like the mode is preferred to avoid over-complicating the model with 'noise' categories.

Interpretation:

This heatmap serves as a health check for the dataset, specifically identifying Multicollinearity—where multiple sensors are effectively telling the model the same thing.

The "Problem Clusters" The Liquid/Sugar Cluster (Bottom Right): There is a massive dark blue block where Density, Balling, Balling_Lvl, Carb_Volume, and Alch_Rel intersect.

Interpretation: These variables are nearly identical (approx 0.95+). In beverage production, density and sugar content are mathematically linked. Including all of them would "confuse" a linear model, making your coefficients unstable.

The Pressure Sync (Top Left): Carb_Pressure1, Fill_Pressure, and Pressure_Setpoint are highly correlated.

Interpretation: The system is perfectly tuned; when the setpoint moves, the sensors follow. We likely only need one of these to represent the "pressure state" of the run.

Brand Artifacts (Red Blocks): The dark red area (e.g., Brand_CodeB vs. Brand_CodeD) is a result of Dummy Encoding.

Interpretation: Since a batch can only be one brand at a time, if it is "Brand B," it mathematically cannot be "Brand D," creating a strong negative correlation.

The (White/Light Zones):

Variables like Temperature, Oxygen_Filler, and Filler_Level are surrounded by white/light colors.

Interpretation: These are Independent Predictors. They aren't twins with anything else, meaning they provide unique, high-value information that the model can't get from any other sensor.

**Removing Redundant Predictors**

```{r}

# Identify the indices of variables to remove based on the 0.90 cutoff
# 'cor_matrix' was created in your previous step
high_cor_indices <- findCorrelation(cor_matrix, cutoff = 0.90)

# Extract the names of the variables being dropped for your report
vars_to_remove <- colnames(cor_matrix)[high_cor_indices]

cat("Predictors identified for removal due to redundancy:\n")
print(vars_to_remove)

# Create the final, optimized datasets
# Removing these columns from both the Training and Evaluation sets
train_ready <- train_data_encoded |> select(-all_of(vars_to_remove))
test_ready  <- test_data_encoded  |> select(-all_of(vars_to_remove))

# Final Sanity Check
cat("\nFinal Preparation Check:\n")
cat("Total Rows in Training:", nrow(train_ready), "\n")
cat("Total Predictors remaining:", ncol(train_ready) - 1, "\n") # -1 for the target PH
cat("Remaining NULLs:", sum(is.na(train_ready)), "\n")

```


```{r message=FALSE, warning=FALSE}
# Check column counts
cat("Train columns (including PH):", ncol(train_ready), "\n")
cat("Test columns (Predictors only):", ncol(test_ready), "\n")

# Check if the predictor names are identical and in the same order
# We remove PH from train for a fair comparison
predictor_match <- all(colnames(train_ready |> select(-PH)) == colnames(test_ready))
cat("Do all predictors match perfectly?", predictor_match)
```

```{r}
#Get the names of the predictors that actually made it into your model
# We remove 'PH' because the test set shouldn't have the answer key
final_predictor_names <- colnames(train_ready |> select(-PH))

# Force the test data to only keep THOSE specific columns
# This automatically drops the 4 extra variables causing the mismatch
test_ready_aligned <- test_ready |> 
  select(all_of(final_predictor_names))

# Final Verification Check
cat("Train predictors:", length(final_predictor_names), "\n")
cat("Test predictors:", ncol(test_ready_aligned), "\n")

# This should now return TRUE
all(colnames(train_ready |> select(-PH)) == colnames(test_ready_aligned))
```



