---
title: "Project_2_Data_624"
author: "Farhod Ibragimov, Alexander Simon, Jackie Yee, Mubashira Qari"
date: "2026-04-18"
output:
  html_document:
    toc: true
    toc_float: true
  word_document:
    toc: true
---

=====================================================
**Loading Datasets**
=====================================================

```{r message=FALSE, warning=FALSE}

# Load libraries
library(tidyverse)
library(mice)      # For advanced imputation
library(caret)     # For data splitting and pre-processing

# Load the datasets
train_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_624_Team_Project_2/refs/heads/main/StudentData.csv", stringsAsFactors = FALSE)

test_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_624_Team_Project_2/refs/heads/main/StudentEvaluation.csv", stringsAsFactors = FALSE)


#view(train_data)
#view(test_data)
```

=====================================================
**Data Exploration (EDA)**
=====================================================

```{r}

```


```{r}

```

=====================================================
**Data Preparation (eg, imputation, transformations)**
=====================================================


```{r}
library(skimr)

skim(train_data)
```


```{r}

# View the structure and summary of the data
str(train_data)
summary(train_data)
```

**Target Variable (PH) Validation**

First we examine the PH column in the training set and remove records that do not have an associated PH value.
If we include rows in our training set where the PH value is missing (NA), the mathematical algorithms will have no answer key to learn from for those specific rows. 

While we can use statistical methods to guess (impute) missing predictor values like temperature, we should never guess the target value in the training data, since this introduces artificial bias into our model. 

Removing these 4 records ensures our model is built only on confirmed, historical production facts.

```{r}
# Count missing values in the target variable
missing_ph <- sum(is.na(train_data$PH))
print(paste("Missing PH values in training set:", missing_ph))

# Remove rows where PH is NA
filtered_data <- train_data |>
  filter(!is.na(PH))

head(filtered_data)
```


```{r}
# Replace all periods with underscores in the column names
colnames(filtered_data) <- gsub("\\.", "_", colnames(filtered_data))
colnames(test_data) <- gsub("\\.", "_", colnames(test_data))

# Quick check to see the new names
names(filtered_data)
```


**Analysis od Dataset**

```{r message=FALSE, warning=FALSE}
# Checking the total number of missing values per column
missing_summary <- colSums(is.na(filtered_data))
print(missing_summary[missing_summary > 0]) # Show only columns with NAs

# Visualizing the missing patterns
# install.packages("VIM")
library(VIM)
aggr(filtered_data, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, 
     labels=names(train_data), cex.axis=.7, gap=3, 
     ylab=c("Histogram of missing data","Pattern"))
```


```{r}
head(filtered_data)
```

In Chapter 3 of Applied Predictive Modeling, the nearZeroVar function uses a strict two-part mathematical test to decide if a predictor is useless (NZV). A predictor is only marked TRUE if it satisfies both of these conditions:

The Frequency Ratio is > 19: This means the most common value appears at least 19 times more often than the second most common value.

The Percent Unique is < 10%: This means the number of unique values in the column is less than 10% of the total number of rows.

**Identify Near-Zero Variance (NZV) Predictors**

```{r}

# 'saveMetrics = TRUE' provides the detailed math behind the decision
nzv_analysis <- nearZeroVar(filtered_data, saveMetrics = TRUE)

nzv_analysis
```
When we look at the raw data for Hyd Pressure1, we can see exactly why it triggered both alarms:

Frequency Ratio (~31.1): The value 0.0 appears 840 times. 
Since 840 / 27 approx. 31.1, it far exceeds the imbalance threshold of 19 and this is below 10%. 
This tells the algorithm that this variable is almost a constant zero.

Consequence of not removing: removing Hyd Pressure1 because if you perform Resampling like Cross-Validation, there is a high statistical chance that in one of folds, that variable will become a constant 0 for every single row. If that happens, many models will crash because they cannot calculate a slope or weight for a column that never changes.


```{r}
# Identify the indices of the predictors to remove

nzv_indices <- nearZeroVar(filtered_data)

# Remove these predictors from the TRAIN and TEST sets
# Saving into NEW variables as requested
train_data_cleaned <- filtered_data[, -nzv_indices]
test_data_cleaned <- test_data[, -nzv_indices]

# 3. Verify the change
# Check how many columns were removed
cat("Original number of columns:", ncol(filtered_data), "\n")
cat("New number of columns:", ncol(train_data_cleaned), "\n")

# View which column was removed
colnames(filtered_data)[nzv_indices]
```


```{r}
```


