---
title: "Project_2_Data_624"
author: "Farhod Ibragimov, Alexander Simon, Jackie Yee, Mubashira Qari"
date: "2026-04-18"
output:
  html_document:
    toc: true
    toc_float: true
  word_document:
    toc: true
---

=====================================================
**Loading Datasets**
=====================================================

```{r message=FALSE, warning=FALSE}

# Load libraries
library(tidyverse)
library(mice)      # For advanced imputation
library(caret)     # For data splitting and pre-processing

# Load the datasets
train_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_624_Team_Project_2/refs/heads/main/StudentData.csv", stringsAsFactors = FALSE)

test_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_624_Team_Project_2/refs/heads/main/StudentEvaluation.csv", stringsAsFactors = FALSE)


#view(train_data)
#view(test_data)
```

=====================================================
**Data Exploration (EDA)**
=====================================================

```{r}

```


```{r}

```

=====================================================
**Data Preparation (eg, imputation, transformations)**
=====================================================


```{r}
library(skimr)

skim(train_data)
```


```{r}

# View the structure and summary of the data
str(train_data)
summary(train_data)
```

**Target Variable (PH) Validation**

First we examine the PH column in the training set and remove records that do not have an associated PH value.
If we include rows in our training set where the PH value is missing (NA), the mathematical algorithms will have no answer key to learn from for those specific rows. 

While we can use statistical methods to guess (impute) missing predictor values like temperature, we should never guess the target value in the training data, since this introduces artificial bias into our model. 

Removing these 4 records ensures our model is built only on confirmed, historical production facts.

```{r}
# Count missing values in the target variable
missing_ph <- sum(is.na(train_data$PH))
print(paste("Missing PH values in training set:", missing_ph))

# Remove rows where PH is NA
filtered_data <- train_data |>
  filter(!is.na(PH))

head(filtered_data)
```

**Column Name Standardization**

```{r}
# Replace all periods with underscores in the column names
colnames(filtered_data) <- gsub("\\.", "_", colnames(filtered_data))
colnames(test_data) <- gsub("\\.", "_", colnames(test_data))

# Quick check to see the new names
names(filtered_data)
```


**Analysis of Dataset**

```{r message=FALSE, warning=FALSE}
# Checking the total number of missing values per column
missing_summary <- colSums(is.na(filtered_data))
print(missing_summary[missing_summary > 0]) # Show only columns with NAs

# Visualizing the missing patterns
# install.packages("VIM")
library(VIM)
aggr(filtered_data, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, 
     labels=names(train_data), cex.axis=.7, gap=3, 
     ylab=c("Histogram of missing data","Pattern"))
```


```{r}
head(filtered_data)
```

In Chapter 3 of Applied Predictive Modeling, the nearZeroVar function uses a strict two-part mathematical test to decide if a predictor is useless (NZV). A predictor is only marked TRUE if it satisfies both of these conditions:

The Frequency Ratio is > 19: This means the most common value appears at least 19 times more often than the second most common value.

The Percent Unique is < 10%: This means the number of unique values in the column is less than 10% of the total number of rows.

**Identify Near-Zero Variance (NZV) Predictors**

```{r}

# 'saveMetrics = TRUE' provides the detailed math behind the decision
nzv_analysis <- nearZeroVar(filtered_data, saveMetrics = TRUE)

nzv_analysis
```
When we look at the raw data for Hyd Pressure1, we can see exactly why it triggered both alarms:

Frequency Ratio (~31.1): The value 0.0 appears 840 times. 
Since 840 / 27 approx. 31.1, it far exceeds the imbalance threshold of 19 and this is below 10%. 
This tells the algorithm that this variable is almost a constant zero.

Consequence of not removing: removing Hyd Pressure1 because if you perform Resampling like Cross-Validation, there is a high statistical chance that in one of folds, that variable will become a constant 0 for every single row. If that happens, many models will crash because they cannot calculate a slope or weight for a column that never changes.


```{r}
# Identify the indices of the predictors to remove

nzv_indices <- nearZeroVar(filtered_data)

# Remove these predictors from the TRAIN and TEST sets
# Saving into NEW variables as requested
train_data_cleaned <- filtered_data[, -nzv_indices]
test_data_cleaned <- test_data[, -nzv_indices]

# 3. Verify the change
# Check how many columns were removed
cat("Original number of columns:", ncol(filtered_data), "\n")
cat("New number of columns:", ncol(train_data_cleaned), "\n")

# View which column was removed
colnames(filtered_data)[nzv_indices]
```


```{r}
# Identifying which column index corresponds to the NZV predictor 
# from our previous nzv_analysis object
nzv_cols <- which(nzv_analysis$nzv == TRUE)

# Removing the columns from train and test datasets

train_data_final <- filtered_data[, -nzv_cols]
test_data_final <- test_data[, -nzv_cols]

# Verifying that the shapes match
dim(train_data_final)
dim(test_data_final)

```

**Transformations to Resolve Skewness**

Many predictive models (especially linear models and Neural Networks) perform much better when the predictors have a symmetric, normal distribution.


**Skewness Visualization**

```{r}
# Load required libraries
library(tidyverse) # Contains dplyr, tidyr, and ggplot2
library(e1071)     # Contains the skewness() function

# Starting the pipeline with your cleaned training data
train_data_final |>
  
  # Keeping only the numeric columns
  select(where(is.numeric)) |>
  
  # Calculating skewness for every remaining column
  summarise(across(everything(), ~skewness(.x, na.rm = TRUE))) |>
  
  # Reshaping the data from wide (one row) to long (two columns: Predictor, Skewness)
  pivot_longer(cols = everything(), names_to = "Predictor", values_to = "Skewness") |>
  
  # Adding the flag to identify which variables are highly skewed
  mutate(Highly_Skewed = ifelse(abs(Skewness) > 1, "High Skew", "Acceptable Skew")) |>
  
  # Pipe the formatted data directly into ggplot
  ggplot(aes(x = reorder(Predictor, Skewness), y = Skewness, fill = Highly_Skewed)) +
  geom_col(color = "black", linewidth = 0.2) +
  coord_flip() + 
  scale_fill_manual(values = c("Acceptable Skew" = "skyblue", "High Skew" = "salmon")) +
  
  # Add the safe threshold lines
  geom_hline(yintercept = c(-1, 1), linetype = "dashed", color = "red", linewidth = 1) +
  
  labs(
    title = "Skewness of All Numeric Predictors",
    subtitle = "Variables extending past the red dashed lines need Box-Cox transformation",
    x = "Predictors",
    y = "Skewness Value"
  ) +
  theme_minimal()
```

The chart identifies which numeric predictors in your dataset have a high skewness extending beyond the safe +/- 1 threshold (red lines), highlighting exactly which variables need a Box-Cox transformation to fix their skewed distributions.

**Centering and Scaling (Individual Predictor Transformations)**

Centering: Subtracting the average (mean) of a predictor from all its values so that the new mean is zero.

Scaling: Dividing each value by the standard deviation so that the new standard deviation is one.

The Scale Problem: In the data, Filler_Speed is around 4,000, while PC_Volume is a small decimal like 0.27.

Without scaling, a model might assume Filler_Speed is 10,000 times more important than PC_Volume just because the numbers are larger. Centering and scaling ensure every variable has the same weight in the model.

Model Requirement: Many algorithms (like KNN, SVM, and Neural Networks) will perform poorly or even fail if the data isn't scaled.

Applying this to the test data:

We must use the Training Data's mean and standard deviation to transform the test data. This ensures the units remain consistent across both sets.

```{r}
# Load the caret library if not already loaded
library(caret)

# calculate 'center' and 'scale' for all numeric columns.
# It will automatically ignore categorical columns like 'Brand_Code' 
# and we exclude the target 'PH' to keep it in its original units.
scaling_plan <- preProcess(train_data_final[, names(train_data_final) != "PH"], 
                           method = c("center", "scale"))

# Apply the transformation to create NEW variables
# This ensures we don't overwrite your 'final' variables
train_data_scaled <- predict(scaling_plan, train_data_final)
test_data_scaled <- predict(scaling_plan, test_data_final)

# Verify the results
# The means should now be 0 (or very close to it, e.g., 10^-16)
# and the standard deviations should be exactly 1.
# We check the first 5 columns as a sample:
print("Means of scaled training data (first 5 columns):")
colMeans(train_data_scaled[, sapply(train_data_scaled, is.numeric)][, 1:5], na.rm = TRUE)

print("Standard Deviations of scaled training data (first 5 columns):")
apply(train_data_scaled[, sapply(train_data_scaled, is.numeric)][, 1:5], 2, sd, na.rm = TRUE)

# Check dimensions to ensure consistency
cat("Train dimensions:", dim(train_data_scaled), "\n")
cat("Test dimensions:", dim(test_data_scaled), "\n")
```

**Scaling Visualization**

```{r message=FALSE, warning=FALSE}
# Load visualization libraries
library(ggplot2)
library(gridExtra)

# --- VISUALIZE BEFORE SCALING ---
# Filler_Speed has a large range (thousands)
p1_before <- ggplot(train_data_final, aes(x = Filler_Speed)) + 
  geom_density(fill = "skyblue", alpha = 0.5) +
  ggtitle("Before: Filler_Speed") +
  xlab("Raw Units (~2500 to 5000)") +
  theme_minimal()

# PC_Volume has a tiny range (decimals)
p2_before <- ggplot(train_data_final, aes(x = PC_Volume)) + 
  geom_density(fill = "salmon", alpha = 0.5) +
  ggtitle("Before: PC_Volume") +
  xlab("Raw Units (~0.1 to 0.5)") +
  theme_minimal()

# --- VISUALIZE AFTER SCALING ---
# Now both will sit on the same X-axis range (roughly -3 to 3)
p1_after <- ggplot(train_data_scaled, aes(x = Filler_Speed)) + 
  geom_density(fill = "skyblue", alpha = 0.5) +
  ggtitle("After: Filler_Speed (Scaled)") +
  xlab("Standard Deviations from Mean") +
  theme_minimal()

p2_after <- ggplot(train_data_scaled, aes(x = PC_Volume)) + 
  geom_density(fill = "salmon", alpha = 0.5) +
  ggtitle("After: PC_Volume (Scaled)") +
  xlab("Standard Deviations from Mean") +
  theme_minimal()

# Arrange all four plots in a 2x2 grid
grid.arrange(p1_before, p2_before, p1_after, p2_after, ncol = 2)
```
Interpretation:

The chart uses two extreme variables—Filler_Speed (huge numbers) and PC_Volume (tiny decimals)—to visually demonstrate why data must be scaled. 

Before scaling, models get tricked into thinking variables with naturally larger numbers are more important. 

By centering the averages to 0 and scaling the spread to 1, both variables are mathematically forced onto the exact same playing field (ranging from roughly -3 to +3). 

While the chart only displays two variables for clarity, this exact same balancing process was applied to all 32 numeric columns so the model can evaluate them fairly.

