---
title: "Project_2_Data_624"
author: "Farhod Ibragimov, Alexander Simon, Jackie Yee, Mubashira Qari"
date: "2026-04-18"
output:
  html_document:
    toc: true
    toc_float: true
  word_document:
    toc: true
---

=====================================================
**Loading Datasets**
=====================================================

```{r message=FALSE, warning=FALSE}

# Load libraries
library(tidyverse)
library(mice)      # For advanced imputation
library(caret)     # For data splitting and pre-processing

# Load the datasets
train_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_624_Team_Project_2/refs/heads/main/StudentData.csv", stringsAsFactors = FALSE)

test_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_624_Team_Project_2/refs/heads/main/StudentEvaluation.csv", stringsAsFactors = FALSE)


#view(train_data)
view(eval_data)
```

=====================================================
**Data Exploration (EDA)**
=====================================================

```{r}

```


```{r}

```

=====================================================
**Data Preparation (eg, imputation, transformations)**
=====================================================


```{r}
library(skimr)

skim(train_data)
```


```{r}

# View the structure and summary of the data
str(train_data)
summary(train_data)
```

**Target Variable (PH) Validation**

First we examine the PH column in the training set and remove records that do not have an associated PH value.
If we include rows in our training set where the PH value is missing (NA), the mathematical algorithms will have no answer key to learn from for those specific rows. 

While we can use statistical methods to guess (impute) missing predictor values like temperature, we should never guess the target value in the training data, since this introduces artificial bias into our model. 

Removing these 4 records ensures our model is built only on confirmed, historical production facts.

```{r}
# Count missing values in the target variable
missing_ph <- sum(is.na(train_data$PH))
print(paste("Missing PH values in training set:", missing_ph))

# Remove rows where PH is NA
filtered_data <- train_data |>
  filter(!is.na(PH))

head(filtered_data)
```


```{r}
# Replace all periods with underscores in the column names
colnames(filtered_data) <- gsub("\\.", "_", colnames(filtered_data))
colnames(test_data) <- gsub("\\.", "_", colnames(test_data))

# Quick check to see the new names
names(filtered_data)
```


**Analysis od Dataset**

```{r message=FALSE, warning=FALSE}
# Checking the total number of missing values per column
missing_summary <- colSums(is.na(train_data))
print(missing_summary[missing_summary > 0]) # Show only columns with NAs

# Visualizing the missing patterns
# install.packages("VIM")
library(VIM)
aggr(train_data, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, 
     labels=names(train_data), cex.axis=.7, gap=3, 
     ylab=c("Histogram of missing data","Pattern"))
```


```{r}
head(filtered_data)
```

In Applied Predictive Modeling by Kuhn & Johnson, the first step is the nearZeroVar function from the caret package.

To identify Near-Zero Variance (NZV) predictors, the book defines two specific thresholds:

Frequency Ratio (freqCut): The ratio of the frequency of the most common value to the frequency of the second most common value. The default threshold is 19 (which represents a 95/5 split).

Percent Unique (uniqueCut): The number of unique values divided by the total number of samples, multiplied by 100. The default threshold is 10%.

A predictor is flagged for removal only if both the Frequency Ratio is greater than 19 and the Percent Unique is less than 10.

**Identify Near-Zero Variance (NZV) Predictors**

```{r}

# 'saveMetrics = TRUE' provides the detailed math behind the decision
nzv_analysis <- nearZeroVar(filtered_data, saveMetrics = TRUE)

nzv_analysis
```

```{r}

```


```{r}
```


