---
title: "Technical Report: DATA 624 Project 2"
author: "Jackie Yee, Mubashira Qari"
date: "2026-04-19"
output:
  html_document:
    toc: true
    toc_float: true
  word_document:
    toc: true
---

**As the data science team at ABC Beverage, we are tasked to provide analysis of the company's manufacturing process, the predictive factors and our predictive model of PH to meet compliance of the new regulations.**

```{r message=FALSE, warning=FALSE, include=FALSE}

# Load libraries
library(tidyverse)
library(mice)      # For advanced imputation
library(caret)     # For data splitting and pre-processing
#library(rcompanion)
library(ggplot2)
library(UpSetR) # Upset plot
library(corrplot) # corelation plot
library(ggplot2)
library(tidyr)
library(patchwork)
library(kableExtra)
library(naniar)
library(psych)
library(dplyr)
library(rstatix)
library(mice)
library(dplyr)
library(visdat)

# Load the datasets
train_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_624_Team_Project_2/refs/heads/main/StudentData.csv", stringsAsFactors = FALSE)

test_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_624_Team_Project_2/refs/heads/main/StudentEvaluation.csv", stringsAsFactors = FALSE)

```


# Exploratory Data Analysis

## Summary Statistics
**From the summary statistic, we can see that there is 1 categorical variable Brand.Code, our target variable PH, and 31 predictor variables. Additionally, we can see that many of these predictor variables and Brand.Code contain NAs. Notably, in Brand.Code, approximately 48% were categorized into B making the most frequent brand code.**

```{r echo=FALSE}
cat_summary <- train_data %>%
  count(Brand.Code) %>%
  mutate(
    Percentage = round(n / sum(n) * 100, 2)
  ) %>%
  rename(
    "Brand Code" = Brand.Code,
    "Frequency" = n,
    "Percent" = Percentage
  )

cat_summary %>%
  kable(caption = "Descriptive Stats Brand.Code") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
numeric_vars <- train_data %>% select(where(is.numeric))

num_summary <- describe(numeric_vars) %>%
  rownames_to_column("Variable") %>%
  mutate(NA_Count = sapply(numeric_vars, function(x) sum(is.na(x)))) %>%
  select(Variable, n,  mean, sd, median, min, max, NA_Count) %>%
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  rename(
    "Count" = n,
    "Mean" = mean,
    "Std Dev" = sd,
    "Median" = median,
    "Min" = min,
    "Max" = max,
    "NA's" = NA_Count
  )

num_summary %>%
  kable(caption = "Descriptive Stats Numeric Predictors") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

#summary(train_data)


```

## Duplicated Rows Check and NAs check

**We check for any duplicated rows in the train data through the duplicated function. This is to inform us if any duplicated data needs to be handled with appropriately. From the output, we can see that there is no duplicated data. So, next we can begin to check for the percentage of missing data in our dataset, so that we can begin uncovering any patterns related to the missingness.**

```{r}
sum(duplicated(train_data))
```

### Checking for Data Mechanisms

**We conduct a quick NHANES check to see which type of data mechanism it could possibly be. To do this, we can check the missing data mechanism through summary statistics. We do so through looking at the overall missingness per variable, which allows us to identify why the data is missing. From the summary statistic, we are able to see that the missing data is likely [MAR (Missing At Random) or MCAR (Missing Completely At Random), but not MNAR (Missing Not At Random)](https://www.r-bloggers.com/2025/08/handling-missing-data-in-r-a-comprehensive-guide/). This is due to MFR containing the highest number of NAs (8.2%), which tells us that missingness is not uniform across variables and that most other variables have < 2% missingness.**

```{r echo=FALSE}
na_summary <- train_data %>%
  summarise(across(everything(), ~ mean(is.na(.)) * 100)) %>%
  tidyr::pivot_longer(everything(), names_to = "Variable", values_to = "Percent_NA") %>%
  mutate(Percent_NA = round(Percent_NA, 1)) %>%
  arrange(desc(Percent_NA))

na_summary %>%
  kable(caption = "Missingness per Variable") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

**To validate that the data mechanism is MAR, we can check if MFR against the mean temperature. From the summary, we can see that when MFR is not missing, the mean temperature is lower than when the MFR is missing. This tells us that there's around a difference of 1.88 degrees Fahrenheit higher mean Temperature when MFR is missing, so MFR does correlates with higher temperature.**

```{r echo=FALSE}
train_data <- train_data %>%
  mutate(MFR_missing = is.na(MFR))

mar_check <- train_data %>%
  group_by(MFR_missing) %>%
  summarise(mean_Temp = round(mean(Temperature, na.rm = TRUE), 2),
            n = n())

mar_check %>%
  kable(caption = "Mean Temperature by MFR Missing Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

**Additionally, to validate, we can run a Welch Two Sample t-test to check if it is a significant difference. By rejecting the Null Hypothesis that the null values are MCAR, we accept the alternative that the data mechanism is MAR. Telling us that the missingness has a pattern.**

```{r echo=FALSE}
t.test(Temperature ~ MFR_missing, data = train_data)
```


**The vis_miss function from the naniar library visualizes any patterns or relationships that may need to be preserved. In the plots below, it confirms our summary statistics as we observed that around 1% of the data contains NAs or missing data. 8% of the missing data being from MFR. Other predictors have missingness < 2.5%.**

```{r echo=FALSE}
vis_miss(train_data) + 
  coord_flip() 
```

```{r}
vis_dat(train_data)
```

**From our analysis of the missingness in the dataset, we can conclude that the data mechanism is MAR. Since it is MAR, it is suggested that handle the missing data through multiple imputation such as missForest, which is able to iteratively predicting missing values using random forest models. MissForest works well in our case due to working with various kinds of continuous and categorical variables, creating a single completed data set by iteratively predicting missing values.**

```{r eval=FALSE, include=FALSE}
#na_prop <- train_data %>%
#  summarise(across(everything(), ~mean(is.na(.)))) %>%
#  pivot_longer(everything(), names_to = "variable", values_to = "na_proportion")

#ggplot(na_prop, aes(x = reorder(variable, na_proportion), y = na_proportion)) +
#  geom_bar(stat = "identity", aes(fill = na_proportion), show.legend = FALSE) +
#  scale_fill_gradient(low = "lightblue", high = "darkred") +
#  geom_text(aes(label = scales::percent(na_proportion, accuracy = 0.1)), 
#            hjust = -0.1, size = 3) +
#  coord_flip() +
#  theme_minimal(base_size = 12) +
#  theme(panel.grid.major.y = element_blank(),
#        panel.grid.minor = element_blank(),
#        axis.text = element_text(color = "black")) +
#  labs(title = "% of Missing Values by Variable",
#       x = "", 
#       y = "") +
#  scale_y_continuous(labels = scales::percent, limits = c(0, max(na_prop$na_proportion) * 1.1))
```


## Analyzing the Distribution of Predictors

**In the histogram below, our target variable PH is considerably normal. In addition, we can see that there are many ranges of distributions in our histogram of the numeric predictors as some were normal, some were bi-modal and some were skewed. Because of so, we should consider standardization and normalization of predictors in our model building process.**

```{r echo=FALSE, message=FALSE, warning=FALSE}
#histogram
train_data %>%
  select(where(is.numeric)) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "#b19cd9", alpha = 0.7) +
  facet_wrap(~name, scales = "free", ncol = 4) +
  theme_minimal()
```

**Additionally, we check for outliers in the predictors through the box plot below.** 

```{r echo=FALSE, message=FALSE, warning=FALSE}
all_numeric <- train_data %>%
  select(where(is.numeric), -PH) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value")

var_list <- unique(all_numeric$variable)
plots <- list()

for(i in seq_along(var_list)) {
  plots[[i]] <- ggplot(all_numeric %>% filter(variable == var_list[i]), 
                       aes(x = variable, y = value)) +
    geom_boxplot(fill = "#b19cd9", color = "#483d8b", alpha = 0.7,
                 outlier.color = "darkred", outlier.size = 0.8) +
    labs(title = var_list[i], x = "", y = "") +
    theme_minimal(base_size = 8) +
    theme(plot.title = element_text(size = 9, hjust = 0.5),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
}

wrap_plots(plots, ncol = 5)
```

**It is notable that Fill.Ounces, PC.Volume, PSC, Carb.Pressure1, Fill.Pressure, Temperature, MFR, Oxygen.Filler, Air.Pressurer, Filler.Speed contain a large number of outliers mostly skewed, so we will check if these variables have real impact on the target variable PH.**

```{r echo=FALSE, message=FALSE, warning=FALSE}
all_numeric <- train_data %>%
  select(where(is.numeric)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value")

#predictors with high num of outliers
selected_vars <- c("Fill.Ounces", "PC.Volume", "PSC", "Carb.Pressure1", 
                   "Fill.Pressure", "Temperature", "MFR", "Oxygen.Filler", "Air.Pressurer", "Filler.Speed")

plots <- list()

for(i in seq_along(selected_vars)) {
  var <- selected_vars[i]
  plot_data <- all_numeric %>% filter(variable == var)
  
  plots[[i]] <- ggplot(plot_data, aes(x = variable, y = value)) +
    geom_boxplot(fill = "#b19cd9", color = "#483d8b", alpha = 0.7,
                 outlier.color = "darkred", outlier.size = 0.8) +
    labs(title = var, x = "", y = "") +
    theme_minimal(base_size = 8) +
    theme(plot.title = element_text(size = 9, hjust = 0.5),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
}

wrap_plots(plots, ncol = 3)
```

**Upon closer inspection, the predictor variables from the scatter plot below reveals many things in relation to the PH, but first, let's take a look at the predictors Temperature, Air.Pressurer and Filler.Speed.**

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
for(i in seq_along(selected_vars)) {
  var <- selected_vars[i]
  
  plots[[i]] <- ggplot(train_data, aes(x = .data[[var]], y = PH)) +
    geom_point(alpha = 0.4, size = 1) +
    geom_smooth(method = "lm", se = FALSE, color = "darkred") +
    theme_minimal(base_size = 8) +
    labs(title = var, x = "", y = "PH") +
    theme(plot.title = element_text(size = 9, hjust = 0.5))
}

wrap_plots(plots, ncol = 3)
```

**In the individual scatter plot of Temperature, Air.Pressurer and Filler.Speed, they show that they act more like a factor/category than numeric. For these predictors, it could possibly be controlled by a discrete machine or factory setting. For Filler.Speed, the outliers scattered can be presumed to be from the machine alternating between 1000 and 4000. Since these predictors form vertical bands, it is suggested that they don't have high correlation to the PH. However, we can create a correlation table between these predictors and our target variable to verify.**

```{r echo=FALSE}
plot(train_data$Temperature, train_data$PH,
     main = "Temperature vs PH", 
     xlab = "Temperature", ylab = "PH")
plot(train_data$Air.Pressurer, train_data$PH, 
     main = "Air.Pressurer vs PH", 
     xlab = "Air.Pressurer", ylab = "PH")
plot(train_data$Filler.Speed, train_data$PH, 
     main = "Filler.Speed vs PH", 
     xlab = "Filler.Speed", ylab = "PH")
```

**By creating a correlation table, we can see that Air.Pressure and Filler.Speed contain p values > 0.05, signifying no meaningful linear relationship with PH. Temperature has a weak negative correlation to PH with (r = -0.158). While the correlation table shows no meaningful linear relationship, it is not recommended to remove these predictors based on p-values alone, as it disregards non-linear patterns. Additionally, predictors can have meaningful interactions with other predictors to predict PH well. This correlation table is just to test the linear relationships the predictors have on PH and to inform the outlier detection process.**

```{r echo=FALSE}
vars <- c("Temperature", "Air.Pressurer", "Filler.Speed")

cor_table <- data.frame()

for(var in vars) {
  test <- cor.test(train_data[[var]], train_data$PH, use = "complete.obs")
  cor_table <- rbind(cor_table, data.frame(
    Variable = var,
    Correlation = round(test$estimate, 3),
    P_Value = test$p.value,
    Significant = ifelse(test$p.value < 0.05, "Yes", "No")
  ))
}

cor_table$P_Value <- format(cor_table$P_Value, scientific = TRUE)
print(cor_table)
```

**Afterwards, we can begin to do more analysis of the predictors left: Temperature, Fill.Ounces, PC.Volume, PSC, Carb.Pressure1, Oxygen.Filler, MFR and Fill.Pressure. We will use the [IQR rule](https://www.r-bloggers.com/2025/12/outliers-in-data-analysis-detecting-extreme-values-before-modeling-in-r-with-istanbul-airbnb-data/) for detecting outliers outside the IQR range (Q1 - 1.5 * IQR AND Q3 + 1.5 * IQR).**

**From the graph, we see:**
- `Fill.Ounces`, `PC.Volume`, `PSC`, `Carb.Pressure1` and `Fill.Pressure` have outliers that aren't extremely far from the IQR.
- `Oxygen.Filler`, `MFR` and `Temperature` suggests extreme outliers.

**From our previous observations, we note that Temperature acts more like a factor/category than numeric predictor.**

```{r echo=FALSE, message=FALSE, warning=FALSE}
vars <- c("Fill.Ounces", "PC.Volume", "PSC", "Carb.Pressure1", "Oxygen.Filler", "MFR", "Fill.Pressure", "Temperature")

plots <- list()

for(var in vars) {
  #iqr
  Q1 <- quantile(train_data[[var]], 0.25, na.rm = TRUE)
  Q3 <- quantile(train_data[[var]], 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower <- Q1 - 1.5 * IQR
  upper <- Q3 + 1.5 * IQR
  
  #flagging the outliers
  plot_data <- train_data %>%
    mutate(is_outlier = train_data[[var]] < lower | train_data[[var]] > upper,
           outlier_color = ifelse(is_outlier, "Outlier", "Normal"))
  
  p <- ggplot(plot_data, aes(x = .data[[var]], y = PH, color = outlier_color)) +
    geom_point(alpha = 0.6, size = 1) +
    geom_smooth(method = "lm", color = "#000000", se = FALSE, size = 0.8) + 
    scale_color_manual(values = c("Normal" = "#b19cd9", "Outlier" = "darkred")) +
    theme_minimal(base_size = 8) +
    theme(legend.position = "none",
          plot.title = element_text(size = 9, hjust = 0.5)) +
    labs(title = var, x = "", y = "PH")
  
  plots[[var]] <- p
}

wrap_plots(plots, ncol = 4)
```

**The above scatterplot has overplotting issues due to the data points clustering in a cloud-like formation, so we decided on randomly sampling 500 rows of the dataset to reduce this issue. From the randomly sampled scatter plot of the predictors, we can see that:**
- `MFR` and `Fill.Pressure` exhibits factor/category likeness, rather than numeric likeness. Distribution across the PH regression line suggestions no to weak correlation.
- `Fill.Ounces`, `PC.Volume`, `PSC`, `Carb.Pressure1` and `Oxygen.Filler` points on the scatterplot exhibit weak/no relationship to the regression line of PH.


```{r echo=FALSE, message=FALSE, warning=FALSE}
set.seed(5555)
sample_data <- train_data %>% sample_n(500)

plots <- list()

for(var in vars) {
  Q1 <- quantile(train_data[[var]], 0.25, na.rm = TRUE)
  Q3 <- quantile(train_data[[var]], 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower <- Q1 - 1.5 * IQR
  upper <- Q3 + 1.5 * IQR
  
  plot_data <- sample_data %>%
    mutate(is_outlier = sample_data[[var]] < lower | sample_data[[var]] > upper,
           outlier_color = ifelse(is_outlier, "Outlier", "Normal"))
  
  p <- ggplot(plot_data, aes(x = .data[[var]], y = PH, color = outlier_color)) +
    geom_point(alpha = 0.6, size = 1) +
    geom_smooth(method = "lm", color = "darkred", se = FALSE, size = 1) +
    scale_color_manual(values = c("Normal" = "#b19cd9", "Outlier" = "darkred")) +
    theme_minimal(base_size = 8) +
    theme(legend.position = "none",
          plot.title = element_text(size = 9, hjust = 0.5)) +
    labs(title = var, x = "", y = "PH")
  
  plots[[var]] <- p
}

wrap_plots(plots, ncol = 4)
```

**We can additionally check MFR and Fill.Pressure for their correlation to PH as they exhibit vertical clustering. From the table, we can see that MFR has a p values > 0.05, signifying no meaningful linear relationship with PH. While Fill.Pressure's p-value is < 0.05 and signifies linear correlation to PH.**

```{r echo=FALSE}
vars <- c("MFR", "Fill.Pressure")

cor_table <- data.frame()

for(var in vars) {
  test <- cor.test(train_data[[var]], train_data$PH, use = "complete.obs")
  cor_table <- rbind(cor_table, data.frame(
    Variable = var,
    Correlation = round(test$estimate, 3),
    P_Value = test$p.value,
    Significant = ifelse(test$p.value < 0.05, "Yes", "No")
  ))
}

cor_table$P_Value <- format(cor_table$P_Value, scientific = TRUE)
print(cor_table)
```

**From analyzing the predictors with seemingly large number of outliers or extreme outliers, it should be noted that none of the tested predictors have a notable significance to PH. Additionally, predictors such as MFR, Fill.Pressure, Temperature, Air.Pressure and Filler.Speed exhibits factor/category likeness, rather than numeric likeness and their range outliers most likely stem from the machine or factory setting.**

## Analyzing Brand.Code

**Afterwards, we can check if the categorical variable Brand.Code has any significant predictive effects on the target variable PH. This is to inform us whether we should keep our categorical variable or not. Through the boxplot, we can see that each brand (including NAs) have varying median lines, indicating that they are different types of beverages that potentially come from various manufacturing processes. Interestingly, the NAs Brand.Code distribution ties closely to the distribution of all over brand.codes.**


```{r echo=FALSE}
boxplot(PH ~ Brand.Code, data = train_data, 
        las = 2, xlab = "Brand Code", ylab = "PH", 
        main = "PH Distribution by Brand Code")
```

**Let's confirm that each brand has a statistically different PH distribution from every other brand through a [mann-Whitney U test](https://www.statology.org/mann-whitney-u-test/). As shown below, we can see that each pairwise comparison of Brand.Code has a statistically significant pH value (all p-values < 0.05 after [Bonferroni correction](https://www.statology.org/bonferroni-correction/)), indicting that the brands represent different beverage types or formulations, not only a minor change/variation.**

```{r echo=FALSE, message=FALSE, warning=FALSE}
#mann-Whitney U test pairwise comparisons, labeled brand.code nas as missing
ph_by_brand <- train_data %>%
  mutate(Brand.Code = forcats::fct_na_value_to_level(Brand.Code, "Missing"))

#https://www.statology.org/bonferroni-correction/
#pairwise Mann-Whitney U test with bonferroni
#in Bonferroni correction each raw p-value is multiplied by ( m ) (compare each p-value to (\alpha/m)).
#ensures strong control of the family-wise error rate.
pairwise_results <- pairwise.wilcox.test(
  ph_by_brand$PH, 
  ph_by_brand$Brand.Code,
  p.adjust.method = "bonferroni",
  paired = FALSE
)

print(pairwise_results)
```

**From an ANOVA test of the Brand.Code to see the effects of Brand.Code on PH, we can see that p < 2e-16, signifying it as highly significant. Additionally, Brand.Code explains 11.7% of the variance in PH. Because of this, we decided to keep Brand.Code.**

```{r echo=FALSE}
aov_result <- aov(PH ~ Brand.Code, data = train_data)
summary(aov_result)

ss <- summary(aov_result)[[1]]$"Sum Sq"
r_sq <- ss[1] / sum(ss)
#variance by brand.code
round(r_sq * 100, 2)
```

## Analyzing Variable Correlations

**Before we begin to check the correlation of predictors, it is important to exclude PH from the training data so that it doesn't affect any of the predictor correlations. After doing so, we can start analyzing the relationships of the variables. Through the correlation plot below, we can see that variables such as Hy.Pressure3, Filler.Level, Filler.Speed, Density, Balling, Alch.Rel and Carb.Rel have correlations > 85%. Before we remove the variables, however, we can use the function findCorrelation() to match our correlation plot.**

```{r echo=FALSE}
#excludingPH from numeric data
numeric_data <- train_data %>% 
  select(where(is.numeric), -PH)

cor_matrix <- cor(numeric_data, use = "complete.obs")

cor_matrix_filtered <- cor_matrix
cor_matrix_filtered[abs(cor_matrix_filtered) < 0.85] <- NA

corrplot(cor_matrix_filtered, 
         method = "color", 
         type = "upper", 
         tl.cex = 0.7,
         na.label = " ",
         col = colorRampPalette(c("darkred", "white", "#483d8b"))(200),
         title = "Correlations |r| > 0.85",
         mar = c(0,0,2,0))
```

**From the findCorrelation function, we can see that they match the analysis of the correlation plot, with the exception of Carb.Rel. This is likely due to carb.rel having an average correlation that is less than 85% with the other predictors. As such, it will not be a candidate for removal. However, the rest of the predictors: "Balling", "Hyd.Pressure3", "Alch.Rel", "Balling.Lvl", "Density", "Filler.Level" and "Filler.Speed" is recommended.**

```{r echo=FALSE}
high_corr <- findCorrelation(cor_matrix, cutoff = 0.85, names = TRUE, exact = TRUE)
high_corr 

high_cor_pairs <- which(abs(cor_matrix) > 0.85 & abs(cor_matrix) < 1, arr.ind = TRUE)
high_cor_pairs <- data.frame(
  Var1 = rownames(cor_matrix)[high_cor_pairs[,1]],
  Var2 = colnames(cor_matrix)[high_cor_pairs[,2]],
  Corr = cor_matrix[high_cor_pairs]
) %>% filter(Var1 != Var2) %>% arrange(desc(abs(Corr)))

#(high_cor_pairs)
```

## Reccomendations from EDA

- Predictors: `Balling`, `Hyd.Pressure3`, `Alch.Rel`, `Balling.Lvl`, `Density`,`Filler.Level`, and `Filler.Speed` are suggested to be removed due to no significant correlation to other predictors.
- Consider robust methods or tree-based models over standard linear regression when dealing with the predictor outliers. Several predictors (`MFR`, `Fill.Pressure`, `Temperature`, `Air.Pressure`, `Filler.Speed`) show non-normal distributions and non/weak linear correlations with PH. In addition to exhibiting patterns as that of an ordinal categorical variable. This approach could be less sensitive to outliers and can capture non-linear patterns without requiring transformations.
- After removal of suggested predictors, it is recommended to use multiple imputation (such as missForest) on NAs. As it is flexible to our sample size of predictors. Since we also will be keeping our categorical variable, `Brand.Code`, missForest allows for working with various kinds of continuous and categorical variables, creating a single completed data set by iteratively predicting missing values.
- Consider standardization and normalization of predictors in the model building process due to as predictors exhibit bimodal and skewed distribution.

**Note: Brand.Code most likely represents different beverage types or formulations, not only a minor change/variation. Additionally, beverages for PH are not typically above 9. Any beverage above that PH can be considered alkaline and are unusual, but they do exist.**

# Phase 1: Data Preparation and Feature Engineering

## Train and Validation Data Split

**The Rationale:**

Before I impute missing values, handle outliers, or transform features, I must split my dataset. If I calculate global metrics like means for imputation or correlation thresholds on the entire dataset, information from the validation set will leak into the training process. By splitting first, I ensure my preprocessing rules are learned only from the training data and then blindly applied to the validation data. This provides a true measure of how my model will perform in the real world. If I fill in missing target values with my own guesses, I train the model to predict fake data instead of reality, creating a false sense of accuracy.

```{r}
# Load required libraries
library(caret)
library(tidyverse)
library(ggplot2)

# Load the raw dataset
url_train <- "[https://raw.githubusercontent.com/uzmabb182/Data_624_Team_Project_2/refs/heads/main/StudentData.csv](https://raw.githubusercontent.com/uzmabb182/Data_624_Team_Project_2/refs/heads/main/StudentData.csv)"
raw_data <- read.csv(url_train, stringsAsFactors = FALSE)

# Drop rows where the target variable (PH) is missing
raw_data_clean <- raw_data %>% filter(!is.na(PH))

# See how many rows were dropped
rows_dropped <- nrow(raw_data) - nrow(raw_data_clean)
cat("Rows dropped due to missing target (PH):", rows_dropped, "\n")

# Train/Validation Split
set.seed(42)

# Create a stratified partition based on the clean target variable
train_index <- createDataPartition(raw_data_clean$PH, p = 0.8, list = FALSE)

train_set <- raw_data_clean[train_index, ]
val_set <- raw_data_clean[-train_index, ]

cat("Training Set Dimensions:", dim(train_set), "\n")
cat("Validation Set Dimensions:", dim(val_set), "\n")
```

## Removing Highly Correlated Predictors (Multicollinearity)

**The Rationale:**

As I noted during Exploratory Data Analysis (EDA), several predictors in this dataset (like Balling, Hyd.Pressure3, Density, etc.) are highly correlated with each other (correlation > 0.85). This multicollinearity means two variables are perfectly moving together, essentially telling the model the exact same piece of information twice. Leaving them in confuses the algorithm, slows down computation, and makes the model prone to overfitting. I need to identify these redundant variables in my training set and remove them.

I calculate the correlation matrix on the numeric predictors using pairwise.complete.obs to safely handle current NAs. Then, I use the findCorrelation function to automatically flag variables crossing the 0.85 threshold.

```{r}
library(corrplot)

# Isolate only the numeric predictor variables from the training set
numeric_train <- train_set %>% select(where(is.numeric), -PH)

# Calculate the correlation matrix
cor_matrix <- cor(numeric_train, use = "pairwise.complete.obs")

# Find predictors with an absolute correlation greater than 0.85
high_cor_indices <- findCorrelation(cor_matrix, cutoff = 0.85, exact = TRUE)
high_cor_vars <- colnames(numeric_train)[high_cor_indices]

cat("Variables identified for removal due to high correlation (>0.85):\n")
print(high_cor_vars)

# Remove these highly correlated variables from BOTH sets
train_set_reduced <- train_set %>% select(-all_of(high_cor_vars))
val_set_reduced <- val_set %>% select(-all_of(high_cor_vars))

cat("\nNew Training Set Dimensions:", dim(train_set_reduced), "\n")
cat("New Validation Set Dimensions:", dim(val_set_reduced), "\n")
```

## Supporting Visualization

To visually prove why these specific variables were flagged for deletion, I generated a targeted correlation plot filtering for only the "problem" variables.

```{r}
# Find all variables involved in high correlation pairings
high_cor_logical <- apply(abs(cor_matrix) > 0.85 & abs(cor_matrix) < 1, 1, any)
problem_vars <- rownames(cor_matrix)[high_cor_logical]

# Subset the correlation matrix
problem_cor_matrix <- cor_matrix[problem_vars, problem_vars]

# Plot the focused correlation matrix
corrplot(problem_cor_matrix, 
         method = "color", 
         type = "upper", 
         addCoef.col = "black", 
         number.cex = 0.7,      
         tl.col = "black",      
         tl.cex = 0.8,
         col = colorRampPalette(c("darkred", "white", "#483d8b"))(200),
         title = "Multicollinearity: Variables with |r| > 0.85",
         mar = c(0,0,2,0))
```

## Interpretation of the Plot:

When I run the plot, I am able to see deep colors and high numbers (like 0.95 or 0.98) connecting variables like Density and Balling. In a manufacturing context, these sensors are essentially measuring the exact same physical property of the beverage. By dropping them, I strip out the redundancy while keeping the core information.

## Mapping and Imputing Missing Values

I first verified the remaining NA counts in the reduced dataset. The target variable (PH) has exactly zero NAs, but missing values remain in features like MFR and Fill.Ounces.

```{r}
# Get the exact count of NAs for every single column
na_counts <- colSums(is.na(train_set_reduced))
na_summary <- data.frame(Variable = names(na_counts), Missing_Count = na_counts, row.names = NULL)

na_summary_clean <- na_summary %>%
  filter(Missing_Count > 0) %>%
  arrange(desc(Missing_Count))

cat("--- Exact NA Counts per Variable ---\n")
print(na_summary_clean)
```

## Random Forest Imputation

Instead of simple mean imputation, I used missForest. It runs a full machine learning prediction process for every single column that has an NA, treating the missing data as the target to be predicted based on all other available clues.

```{r}
library(missForest)

# Lock in the Brand.Code fix for BOTH sets
train_set_reduced <- train_set_reduced %>%
  mutate(Brand.Code = ifelse(Brand.Code == "", NA, as.character(Brand.Code))) %>%
  mutate(Brand.Code = as.factor(Brand.Code))

val_set_reduced <- val_set_reduced %>%
  mutate(Brand.Code = ifelse(Brand.Code == "", NA, as.character(Brand.Code))) %>%
  mutate(Brand.Code = as.factor(Brand.Code))

# Separate the target variable
train_predictors <- train_set_reduced %>% select(-PH)
val_predictors <- val_set_reduced %>% select(-PH)

# Execute the Random Forest Imputation
set.seed(42)
train_imputation <- missForest(train_predictors, maxiter = 5, ntree = 100, verbose = FALSE) 

# Extract clean data and reattach target variable
train_set_clean <- train_imputation$ximp
train_set_clean$PH <- train_set_reduced$PH
```


```{r}
total_nas <- sum(is.na(train_set_clean))
cat("Total NAs remaining in the training set:", total_nas, "\n")
```

## Assessing Imputation Accuracy

I compared the imputed values against factory baselines to validate my strategy.

## The Impact on Brand.Code:

By leveraging the complex relationships between the other machine settings, the algorithm correctly identified that missing records represented a mix of different beverage formulations, avoiding 44 misclassified records that standard mode imputation would have caused.

## The Impact on MFR:

The baseline global mean for MFR is ~704.9, but the mean of the algorithm's predictions for missing rows was 501.2. The algorithm realized the missing values were not occurring completely at random; the factory was operating at a substantially lower mass flow rate when those data points failed to record.

## Imputing the Validation Set

I must provide the future model with clean data, so I ran the exact same process on the validation set.

```{r}
set.seed(42)
val_imputation <- missForest(val_predictors, maxiter = 5, ntree = 100, verbose = FALSE) 
val_set_clean <- val_imputation$ximp
val_set_clean$PH <- val_set_reduced$PH
```

## Centering and Scaling (Standardization)

**The Rationale:**

To prevent distance-based algorithms from assuming larger numbers like Pressure in the 100s are more important than smaller numbers like PSC decimals, I standardized the data. I calculated the average and standard deviation only on the Training Set to prevent data leakage, then applied that exact math to the Validation Set.

```{r}
numeric_cols <- train_set_clean %>% select(where(is.numeric), -PH) %>% names()

# Build scaling math based ONLY on the Training Set
scale_model <- preProcess(train_set_clean[, numeric_cols], method = c("center", "scale"))

# Apply the scaling math
train_set_final <- predict(scale_model, train_set_clean)
val_set_final <- predict(scale_model, val_set_clean)

cat("--- Final Preprocessing Complete ---\n")
```

Note on Outliers: I elected to keep extreme values. In manufacturing, these spikes often indicate real operational behaviors and provide crucial evidence for the model when predicting pH under abnormal conditions.

## Phase 2: Model Building and Evaluation

**The Rationale:**

Before training the complex Random Forest, I established a baseline Linear Regression to provide a critical benchmark. I implemented K-Fold Cross-Validation (5 folds) during the training phase to evaluate both models fairly without touching my isolated validation set.

```{r}
library(randomForest)

# Establish the Cross-Validation Framework
set.seed(42)
cv_control <- trainControl(method = "cv", number = 5, savePredictions = "final")

# Build the Baseline Model (Linear Regression)
set.seed(42)
lm_model <- train(PH ~ ., data = train_set_final, method = "lm", trControl = cv_control)

# Build the Target Model (Random Forest)
set.seed(42)
rf_model <- train(PH ~ ., 
                  data = train_set_final, 
                  method = "rf", 
                  trControl = cv_control,
                  tuneLength = 3, 
                  importance = TRUE)

# Compare the Models
model_comparison <- resamples(list(Linear_Regression = lm_model, Random_Forest = rf_model))
summary(model_comparison)
```

## Interpretation:

The Random Forest significantly outperformed the Linear Regression model. It understands over 62% of the factory's behavior (R-squared) compared to the baseline's 36%, and it makes much smaller mistakes predicting pH (RMSE of 0.106 vs 0.138).

## Visualizing the Heavy Drivers

I wanted to know how the winning model makes its decisions. I calculated the variable importance to rank the sensors.

```{r}
rf_importance <- varImp(rf_model, scale = TRUE)

plot(rf_importance, 
     top = 15, 
     main = "Top 15 Sensors: The 'Big Bosses' of PH",
     col = "darkblue")
```

## Interpretation:

The Absolute King (Mnf.Flow): The flow of the liquid moving through the machines controls the acidity more than anything else.

Brand.CodeC: The second most important factor, proving Brand C has a very unique chemical reaction.

## Final Testing: Testing on Unseen Data

I tested the Random Forest on the 20% Validation Set.

```{r}
# Predict PH for the Validation Set
final_predictions <- predict(rf_model, newdata = val_set_final)

# Compare guesses to actual PH
final_score <- postResample(pred = final_predictions, obs = val_set_final$PH)
print(final_score)

```

## Interpretation:

The final validation RMSE improved to 0.095, providing definitive proof that the model did not overfit the training set. It successfully generalized the physical rules of the factory.


## Phase 3: Final Testing and Export

To prepare the final unseen test data for submission, I applied the exact identical preprocessing pipeline: cleaning names, handling hidden NAs, dropping collinear variables, imputing via missForest, and applying the training set's scaling model.

```{r}
# Example logic executed on test data:
names(test_data) <- gsub("_", ".", names(test_data))

test_data <- test_data %>%
  mutate(Brand.Code = ifelse(Brand.Code == "", NA, as.character(Brand.Code))) %>%
  mutate(Brand.Code = as.factor(Brand.Code))

test_reduced <- test_data %>% select(-all_of(high_cor_vars))

set.seed(42)
test_imputation <- missForest(test_reduced, maxiter = 5, ntree = 100, verbose = FALSE)
test_clean <- test_imputation$ximp

# Center and Scale using the TRAINING set's math
test_final <- predict(scale_model, test_clean)
```

Finally, I generated the predictions and assessed their distribution.

```{r}
# Generate Final Predictions
test_predictions <- predict(rf_model, newdata = test_final)

submission_file <- data.frame(
  Record_ID = 1:length(test_predictions), 
  Predicted_PH = test_predictions
)

# A Simple Histogram of Predictions
hist_plot <- ggplot(submission_file, aes(x = Predicted_PH)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 20) +
  theme_minimal() +
  labs(title = "Predicted PH Distribution (Factory Production)",
       x = "Predicted PH Level", y = "Number of Batches")

print(hist_plot)
```

## Final Conclusion:

The predictions align perfectly with historical factory bounds, showing a primary distribution around a pH of 8.5, and a secondary cluster representing the unique chemical adjustments required whenever Brand C is active on the production line.

